---
title: "When Environment, Health, and Wealth Don’t Align"
author: 
date: 
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
    toc: true
    navbar_fixed: false
    theme:
      version: 4
      bootswatch: flatly
      primary: "#1a1818"
      secondary: "#18234a"
      base_font:
        google: "Source Sans Pro"
      heading_font:
        google: "Montserrat"
---

```{css} 
/* navbar styling */
.navbar {
  min-height: 90px;
  padding-top: 15px;
  padding-bottom: 15px;
  font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  background-color: #16255b;
}

/* navigation links */
.navbar-nav > li > a {
  font-size: 18px;
  padding-left: 14px;
  padding-right: 14px;
}

/* active tab */
.navbar-nav > .active > a {
  color: #cc0000 !important;
  font-weight: bold;
}

/* overall dashboard text */
body {
  font-size: 18px;
  line-height: 1.7;
}

/* paragraph/body text */
p {
  font-size: 20px;
  line-height: 1.8;
}

/* dashboard text inside panels */
.chart-stage p {
  font-size: 20px;
  line-height: 1.8;
}

/* text inside dashboard panels */
.chart-wrapper {
  font-size: 18px;
}

/* section headers */
.section.level2 h2 {
  font-size: 28px;
  font-weight: bold;
}

/* subsection headers */
.section.level3 h3 {
  font-size: 20px;
  font-weight: 600;
}

/* chart titles */
.chart-title {
  font-size: 22px !important;
  font-weight: bold;
}

/* sidebar styling */
.sidebar {
  font-size: 18px;
  line-height: 1.7;
  background-color: #f7f3f4;
  padding: 20px;
}

/* allow scrolling tables */
.chart-wrapper {
  overflow-y: auto;
}

/* widen dashboard */
.main-container {
  max-width: 1700px;
}
```

```{r setup, include=FALSE}

#Note: This runs first and prepares the environment.
#Note: We suppress warnings/messages for a clean dashboard.

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

#Note: Core libraries for data manipulation and visualization
library(tidyverse)
library(plotly)
library(flexdashboard)
library(broom)
library(DT)
library(leaflet)
library(readr)
library(dplyr)
library(sf)
library(ggplot2)
library(janitor)
library(scales)
library(ggrepel)
library(ggcorrplot)
library(patchwork)


#Note: SUNSET COLOR PALETTE
#Note: Consistent colors across all visuals improves readability and professionalism
sun_orange <- "#FF7A00"
sun_yellow <- "#FFC300"
sun_red    <- "#FF3B3B"
sun_pink   <- "#FF8FA3"
dark_gray  <- "#2F2F2F"

# load data
url <- "https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/health_data.csv"
health_data2 <- read.csv(url)

# read shapefile of Chicago
chicago_sf <- st_read("https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/chi_comm_areas.geojson", quiet = TRUE)

# join health data with shapefile
chicago_map <- chicago_sf %>%
  left_join(health_data2 %>%
              mutate(GEOID = as.character(GEOID)), by = c("area_numbe" = "GEOID"))

# pull hardship index from most recent census data from github repository
url2 <- "https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/h_index.csv"
h_index <- read.csv(url2) %>%
  slice(-1, ) # remove first row

# join hardship index to health + map data
chicago_map <- chicago_map %>%
  left_join(h_index %>% select(GEOID, `HDX_2020.2024`) %>%
              mutate(GEOID = as.character(GEOID), `HDX_2020.2024` = as.numeric(`HDX_2020.2024`)), by = c("area_numbe" = "GEOID"))

# transform data into percents
chicago_map <- chicago_map %>%
  mutate(
    diabetes_rate = HCSDIA_2023.2024 / Population * 100,
    obesity_rate = HCSOB_2023.2024 / Population * 100,
    hypertension_rate = HCSHYT_2023.2024 / Population * 100,
    asthma_rate = HCSATH_2023.2024 / Population * 100,
    hard_index = HDX_2020.2024,
    pct_white = PCT.W_2020.2024,
    disease_burden = rowMeans(cbind(diabetes_rate, obesity_rate, hypertension_rate, asthma_rate), na.rm = TRUE)
  ) %>% # add resilience and vulnerability scores
  mutate(
    resilient = hard_index > quantile(hard_index, 0.5, na.rm = TRUE) &
                disease_burden < quantile(disease_burden, 0.5, na.rm = TRUE),
    vulnerable = hard_index < quantile(hard_index, 0.5, na.rm = TRUE) &
                 disease_burden > quantile(disease_burden, 0.5, na.rm = TRUE)
  ) %>% # add obesity specific resilience and vulnerability scores
  mutate(
    resilient_ob = hard_index > quantile(hard_index, 0.5, na.rm = TRUE) &
                    obesity_rate < quantile(obesity_rate, 0.5, na.rm = TRUE),
    vulnerable_ob = hard_index < quantile(hard_index, 0.5, na.rm = TRUE) &
                     obesity_rate > quantile(obesity_rate, 0.5, na.rm = TRUE),
    category_ob = case_when(
      resilient_ob == TRUE ~ "Resilient",
      vulnerable_ob == TRUE ~ "Vulnerable",
      TRUE ~ "Expected"
    )
  )

# remove Fuller Park and Burnside - raw number data do not match CHI Atlas rates, suspected data quality issue
chicago_map <- chicago_map %>%
  filter(!community %in% c("FULLER PARK", "BURNSIDE"))

# set up cor data
cor_data <- chicago_map %>%
  st_drop_geometry() %>%
  select(
    Diabetes = diabetes_rate,
    Obesity = obesity_rate,
    Hypertension = hypertension_rate,
    Asthma = asthma_rate,
    Pct_White = PCT.W_2020.2024,
    Air_Quality = CHASBQJ_2023.2024,
    Traffic_Risk = TRF_2020,
    Env_Justice = CHAKNKC_2023,
    Hardship = hard_index,
    Community = Name
  ) %>%
  drop_na()
```

Overview
=====================================

Sidebar {.sidebar}
-------------------------------------

### Dashboard Navigation

- Overall Disease Trends
- Pollution Overview
- Obesity
- Hypertension
- Asthma
- Diabetes
- Conclusions
- Acknowledgements

### Project Theme

This dashboard explores where environmental exposure, chronic disease burden, and structural inequality fail to align across Chicago neighborhoods.

### Key Variables

- PM2.5
- NO2
- Asthma
- Hypertension
- Diabetes
- Obesity

## Row

### Introduction

Urban environmental conditions are rarely distributed equally across cities in the United States. In Chicago, air pollution exposure can vary substantially between neighborhoods due to traffic density, industrial infrastructure, and population density. At the same time, biological and social indicators such as poverty, chronic illness, and health vulnerability are also unevenly distributed across census tracts.

This project combines two independent data sets, environmental sensor data and Chicago Health Atlas biological indicators, to investigate whether these patterns overlap in a meaningful way. Rather than simply searching for correlation, the purpose of this analysis is to determine whether the observed relationship reflects a realistic ecological pattern connected to how communities experience environmental stressors.

A major challenge in this project is that the data sets operate at different scales and timelines. The environmental data set contains long-term sensor observations collected continuously over time, while the Chicago Health Atlas provides tract-level snapshot indicators collected during a different period. Because of this mismatch, the analysis focuses on creating standardized summary measurements that allow the data sets to be compared responsibly.

### *The central research question guiding this project is:*

**Do Chicago neighborhoods with higher socioeconomic vulnerability also experience greater exposure to air pollution?**

To answer this question, this analyses focuses on the relationship between:

Average black carbon concentrations (environmental burden)
Poverty rate by census tract (biological and social vulnerability)

This relationship was chosen because it reflects a biologically and ecologically meaningful urban pattern. Black carbon pollution is commonly associated with vehicle emissions and combustion sources, while poverty can influence housing location, environmental exposure, healthcare access, and long-term health vulnerability. If both variables cluster together spatially, the relationship may suggest broader environmental inequality patterns rather than random coincidence.

## Row

### **Purpose of the Project**

The purpose of this project is not only to identify a statistical relationship, but also to evaluate whether the relationship:

- Reflects realistic neighborhood-level conditions
- Remains consistent across multiple visualizations
- Demonstrates ecological credibility
- Avoids misleading conclusions caused by unrelated variables.


### **Main Goal**

**The goals of this project was to:**

- Identify relationships between environmental and biological variables
- Determine whether the relationship is statistically meaningful
- Show that the relationship makes ecological and biological sense
- Avoid false positives or misleading associations
 

Overall Disease Trends
=====================================

## Row {data-height="400"}

### Disease Burden by Community Area

```{r disease maps, echo=FALSE, out.width="100%", results='asis'}
# create chloropleth function
make_map <- function(var, title, midpt) {
  ggplot(chicago_map) +
    geom_sf(aes(fill = .data[[var]],
                text = paste0("<b>", Name, "</b><br>",
                              title, ": ", round(.data[[var]], 1), "%"))) +
    scale_fill_gradient2(low = "#2e4f4f",
                         mid = "#f4bb8f",
                         high = "#ff0000",
                         midpoint = midpt,
                         na.value = "black",
                         name = "% Adults") +
    theme_void() +
    theme(legend.position = "none")
}

# construct maps
pd <- make_map("diabetes_rate", "Diabetes", mean(chicago_map$diabetes_rate, na.rm = TRUE))
po <- make_map("obesity_rate", "Obesity", mean(chicago_map$obesity_rate, na.rm = TRUE))
ph <- make_map("hypertension_rate", "Hypertension", mean(chicago_map$hypertension_rate, na.rm = TRUE))
pa <- make_map("asthma_rate", "Asthma", mean(chicago_map$asthma_rate, na.rm = TRUE))

# generate interactive maps of spatial patterns of disease
pd2 <- ggplotly(pd, tooltip = "text") %>% layout(autosize = TRUE)
po2 <- ggplotly(po, tooltip = "text") %>% layout(autosize = TRUE)
ph2 <- ggplotly(ph, tooltip = "text") %>% layout(autosize = TRUE)
pa2 <- ggplotly(pa, tooltip = "text") %>% layout(autosize = TRUE)

# arrange plots in a line
subplot(pd2, po2, ph2, pa2, nrows = 1) %>%
  layout(
    autosize = TRUE,
    annotations = list(
      list(x = 0.125, y = 1.0, text = "<b>Diabetes</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center",
           font = list(size = 14, color = "#7b4419")),
      list(x = 0.375, y = 1.0, text = "<b>Obesity</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center",
           font = list(size = 14, color = "#7b4419")),
      list(x = 0.625, y = 1.0, text = "<b>Hypertension</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center",
           font = list(size = 14, color = "#7b4419")),
      list(x = 0.875, y = 1.0, text = "<b>Asthma</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center",
           font = list(size = 14, color = "#7b4419"))
    )
  ) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Mean Diabetes Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$diabetes_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Diabetes Rate",
  icon = "fa-droplet",
  color = "#2d4f4f"
)
```

### Mean Obesity Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$obesity_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Obesity Rate",
  icon = "fa-weight-hanging",
  color = "#f4bb8f"
)
```

### Mean Hypertension Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$hypertension_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Hypertension Rate",
  icon = "fa-heart-pulse",
  color = "#2d4f4f"
)
```

### Mean Asthma Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$asthma_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Asthma Rate",
  icon = "fa-lungs",
  color = "#f4bb8f"
)
```

## Row {data-height="600"}

### Summary Table of Disease Burden {data-width=400}

```{r top diabetes neighborhoods}
chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(diabetes_rate)) %>%
  arrange(desc(diabetes_rate)) %>%
  select(Community = community, Diabetes = diabetes_rate, Obesity = obesity_rate, Hypertension = hypertension_rate, Asthma = asthma_rate) %>%
  mutate(across(where(is.numeric), ~round(., 1))) %>%
  DT::datatable(options = list(
  pageLength = 67,
  dom = 't',
  scrollY = "450px",
  scrollCollapse = TRUE
))
```

### Spatial Determinants of Disease Burden 

``` {r chloropleth disease burden }
# calculate mean of all disease rates
chicago_map <- chicago_map %>%
  mutate(disease_burden = rowMeans(cbind(diabetes_rate, obesity_rate, hypertension_rate, asthma_rate), na.rm = TRUE))

# visualize on map
disease_burden <- ggplot(chicago_map) +
  geom_sf(aes(fill = disease_burden,
              text = paste0("<b>", Name, "</b><br>",
              "<b>Disease Burden:</b> ", round(disease_burden, 1), "%"))) +
  scale_fill_gradient2(low = "#2e4f4f", mid = "#f4bb8f", high = "#ef0300",
                       midpoint = mean(chicago_map$disease_burden, na.rm = TRUE),
                       na.value = "black", name = NA) +
  theme_void() +
  theme(plot.title = element_text(size = 14, color = "#7b4419", face = "bold", hjust = 0.5)) +
  labs(title = "Overall Disease Burden by Community Area")

ggplotly(disease_burden, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Highest Overall Burden

```{r}
top_burden <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(disease_burden)) %>%
  slice_max(disease_burden, n = 1)

valueBox(
  value = paste0(round(top_burden$disease_burden, 1), "%"),
  caption = paste0("Highest Overall Burden — ", top_burden$community),
  icon = "fa-location-dot",
  color = "#7b4419"
)
```

### Lowest Overall Burden

```{r}
bot_burden <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(disease_burden)) %>%
  slice_min(disease_burden, n = 1)

valueBox(
  value = paste0(round(bot_burden$disease_burden, 1), "%"),
  caption = paste0("Lowest Overall Burden — ", bot_burden$community),
  icon = "fa-location-dot",
  color = "#48bea1"
)
```

### Highest Hardship Index

```{r}
top_hardship <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(hard_index)) %>%
  slice_max(hard_index, n = 1)

valueBox(
  value = round(top_hardship$hard_index, 1),
  caption = paste0("Highest Hardship — ", top_hardship$community),
  icon = "fa-location-dot",
  color = "#ed4a1a"
)
```

### Lowest Hardship Index

```{r}
bot_hardship <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(hard_index)) %>%
  slice_min(hard_index, n = 1)

valueBox(
  value = round(bot_hardship$hard_index, 1),
  caption = paste0("Lowest Hardship — ", bot_hardship$community),
  icon = "fa-location-dot",
  color = "#167b2b"
)
```
## Row {data-height="450"}

### Racial Determinants of Disease Burden and Hardship

``` {r scatterplot disease burden vs hardship with percent white}
scatter <- ggplot(chicago_map, aes(x = hard_index, y = disease_burden, 
                                    color = pct_white,
                                    text = paste0("<b>", community, "</b><br>",
                                                  "Hardship: ", round(hard_index, 1), "<br>",
                                                  "Disease Burden: ", round(disease_burden, 1), "%<br>",
                                                  "% White: ", round(PCT.W_2020.2024, 1), "%"))) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", color = "black") +
  scale_color_gradient(low = "#753D12", high = "#ffffff", name = "% White") +
  labs(x = "Hardship Index", y = "Disease Burden (%)") +
  theme_minimal()

ggplotly(scatter, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

### Correlation Matrix

```{r corr plot}
# create a correlation matrix of all variables
cor_matrix <- cor(cor_data %>%
                    select(-Community)
                  )

p_cor <- ggcorrplot(cor_matrix,
           method = "circle",
           type = "lower",
           lab = FALSE,
           colors = c("red2", "white", "darkslategray"),
           title = "Correlation Matrix") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_cor) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Correlation: % White & Disease Burden

```{r}
cor_val <- cor(chicago_map$PCT.W_2020.2024, chicago_map$disease_burden, use = "complete.obs")

valueBox(
  value = round(cor_val, 2),
  caption = "Correlation: % White & Disease Burden",
  icon = "fa-chart-line",
  color = "#2d4f4f"
)
```

### Correlation: % White & Hardship

```{r}
cor_val2 <- cor(chicago_map$PCT.W_2020.2024, chicago_map$hard_index, use = "complete.obs")

valueBox(
  value = round(cor_val2, 2),
  caption = "Correlation: % White & Hardship Index",
  icon = "fa-chart-line",
  color = "#ed4a1a"
)
```

### Correlation: Hardship & Disease Burden

```{r}
cor_val3 <- cor(chicago_map$hard_index, chicago_map$disease_burden, use = "complete.obs")

valueBox(
  value = round(cor_val3, 2),
  caption = "Correlation: Hardship & Disease Burden",
  icon = "fa-chart-line",
  color = "#f4bb8f"
)
```

### Number of Neighborhoods Above Trend

```{r}
model_burden <- lm(disease_burden ~ hard_index, data = chicago_map)
above <- sum(residuals(model_burden) > 0, na.rm = TRUE)

valueBox(
  value = above,
  caption = "Neighborhoods Above Expected Burden",
  icon = "fa-triangle-exclamation",
  color = "#7b4419"
)
```

## Row {data-height="600"}

### Intra-Categorical Correlation Plot: Disease

``` {r intracat corr disease plot}
# corrplot disease
cor_matrix_disease <- cor(cor_data %>%
                            select(Diabetes, Obesity, Hypertension, Asthma)
                          )

#plot with ggcorrplot
p_cor_disease <- ggcorrplot(cor_matrix_disease,
           method = "circle",
           type = "lower",
           lab = FALSE,
           colors = c("red2", "white", "darkslategray")) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

# ggplotly it
ggplotly(p_cor_disease) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)

```

### Disease vs. Environmental Factors

``` {r disease v environment}

# create vectors of two categories
sub_disease <- c("Diabetes", "Obesity", "Hypertension", "Asthma")
sub_environ <- c("Pct_White", "Air_Quality", "Traffic_Risk", "Env_Justice", "Hardship")

cor_matrix2 <- cor(cor_data %>%
                    select(all_of(c(sub_disease, sub_environ))),
                  use = "complete.obs")

p_cor_matrix2 <- ggcorrplot(cor_matrix2[sub_disease, sub_environ],
           colors = c("darkslategray", "white", "red2")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_line(color = "white", linewidth = 1))

ggplotly(p_cor_matrix2) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

### Intra-Categorical Correlation Plot: Social Determinants of Health

``` {r intra corr environmental factors plot}
cor_matrix_sdoh <- cor(cor_data %>%
                            select(Pct_White, Air_Quality, Traffic_Risk, Env_Justice, Hardship)
                          )

p_cor_sdoh <- ggcorrplot(cor_matrix_sdoh,
           method = "circle",
           type = "lower",
           lab = FALSE,
           lab_size = 3,
           colors = c("red2", "white", "darkslategray")) +
  scale_x_discrete(limits = rev) +
  scale_y_discrete(position = "right") +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_cor_sdoh) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)

```

Pollution Overview
=====================================
## Row {data-height="600"}

### Pollution Overview

```{r pollution-overview-panel, echo=FALSE, warning=FALSE, message=FALSE, fig.width=13, fig.height=8}

# Loaded libraries for data filtering and maps.
library(tidyverse)
library(sf)
library(patchwork)

#Note: Used pollution sensor dataset and Chicago community boundaries.
data_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Open_Air_Chicago_Day_Aggregations_20260212(1)%20(3).csv"
community_url <- "https://data.cityofchicago.org/resource/igwz-8jzy.geojson"

pollution <- read_csv(data_url)

# Imported Chicago neighborhood map and converted coordinates.
chicago_areas <- st_read(community_url, quiet = TRUE) %>%
  st_transform(4326) %>%
  mutate(area_numbe = as.numeric(area_numbe))

#Note: Cleaned pollution sensor points and calculate average pollutant values.
points <- pollution %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  group_by(sensor_name, latitude, longitude) %>%
  summarise(
    pm25 = mean(pm2_5ConcMass24HourMean.value, na.rm = TRUE),
    no2 = mean(no2Conc24HourMean.value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  #Note: Used standardized pollution values and create a combined burden score.
  mutate(
    pm25_z = as.numeric(scale(pm25)),
    no2_z = as.numeric(scale(no2)),
    burden = pm25_z + no2_z
  ) %>%
  
  #Note: Convert pollution points into spatial features
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Match pollution sensor points to Chicago community areas
joined <- st_join(points, chicago_areas, join = st_within)

#Note: Calculated the average pollution burden for each neighborhood.
area_pollution <- joined %>%
  st_drop_geometry() %>%
  group_by(area_numbe) %>%
  summarise(
    pm25 = mean(pm25, na.rm = TRUE),
    no2 = mean(no2, na.rm = TRUE),
    burden = mean(burden, na.rm = TRUE),
    .groups = "drop"
  )

#Note: Combine pollution summaries with neighborhood boundaries
pollution_map <- chicago_areas %>%
  left_join(area_pollution, by = "area_numbe")

#Note: Created a shared theme for all three maps.
map_theme <- theme_void() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 16,
      color = "#7a3f16"
    ),
    legend.position = "bottom",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

#Note: Created first map displaying PM2.5 concentration across Chicago neighborhoods.
p1 <- ggplot(pollution_map) +
  geom_sf(aes(fill = pm25), color = "gray35", linewidth = 0.2) +
  scale_fill_gradient(
    low = "#f8c8dc",
    high = "#7a003c",
    na.value = "gray75",
    name = "PM2.5"
  ) +
  labs(title = "PM2.5") +
  map_theme

#Note: Created second map displaying NO2 concentration across Chicago neighborhoods
p2 <- ggplot(pollution_map) +
  geom_sf(aes(fill = no2), color = "gray35", linewidth = 0.2) +
  scale_fill_gradient(
    low = "#f8c8dc",
    high = "#7a003c",
    na.value = "gray75",
    name = "NO2"
  ) +
  labs(title = "NO2") +
  map_theme

#Note: Created third map by combining the pollution burden score.
p3 <- ggplot(pollution_map) +
  geom_sf(aes(fill = burden), color = "gray35", linewidth = 0.2) +
  scale_fill_gradient(
    low = "#f8c8dc",
    high = "#7a003c",
    na.value = "gray75",
    name = "Burden"
  ) +
  labs(title = "Overall Pollution Burden") +
  map_theme

#Note: Arranged the three pollution maps into one dashboard panel.
(p1 | p2 | p3) +
  plot_annotation(
    title = "Air Pollution Across Chicago Community Areas",
    subtitle = "PM2.5, NO2, and combined pollution burden from Open Air Chicago sensors",
    theme = theme(
      plot.title = element_text(
        hjust = 0.5,
        face = "bold",
        size = 22
      ),
      plot.subtitle = element_text(
        hjust = 0.5,
        size = 12
      )
    )
  )
```


## Row {data-height="450"}

### Pollution Hotspots across Chicago
```{r pollution-hexbin-map, echo=FALSE, warning=FALSE, message=FALSE}

#Note: Loaded tidyverse for data cleaning and plotting.
library(tidyverse)

#Note: Cleaned pollution dataset and removed missing coordinates.
hex_data <- pollution %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  
  #Note: Standardized PM2.5 and NO2 values into z-scores.
  mutate(
    pm25_z = as.numeric(scale(pm2_5ConcMass24HourMean.value)),
    no2_z = as.numeric(scale(no2Conc24HourMean.value)),
    
    #Note: Combined pollutants into one overall pollution burden score.
    pollution_burden = pm25_z + no2_z
  ) %>%
  
  #Note: Removed rows with missing burden values.
  filter(!is.na(pollution_burden))

#Note: Created hexbin map of pollution hotspots across Chicago.
ggplot(hex_data, aes(x = longitude, y = latitude, z = pollution_burden)) +
  
  #Note: Aggregated nearby pollution points into hexagonal bins.
  stat_summary_hex(
    aes(fill = after_stat(value)),
    bins = 25,
    fun = mean
  ) +
  
  #Note: Applied pink-to-dark-purple pollution gradient.
  scale_fill_gradient(
    low = "#f8c8dc",
    high = "#7a003c"
  ) +
  
  #Note: Added all axes labels and titles.
  labs(
    title = "Pollution Hotspots Across Chicago",
    subtitle = "Hexbins show average combined PM2.5 and NO2 burden",
    x = "Longitude",
    y = "Latitude",
    fill = "Pollution Burden"
  ) +
  
  #Note: Kept minimal clean theme.
  theme_minimal() +
  
  #Note: Customized title and subtitle appearance.
  theme(
    plot.title = element_text(
      size = 22,
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 12,
      hjust = 0.5
    )
  )
```




### Pollution Hotspots across Chicago
```{r bivariate-pollution-map, echo=FALSE, warning=FALSE, message=FALSE}

#Note: Loaded libraries for data cleaning and spatial mapping.
library(tidyverse)
library(sf)

#Note: Created pollution burden categories using PM2.5 and NO2 levels.
bivar_map <- pollution_map %>%
  mutate(
    pm25_level = ntile(pm25, 3),
    no2_level = ntile(no2, 3),
    bivar_class = paste0(pm25_level, "-", no2_level)
  )

#Note: Created custom color palette for bivariate pollution categories.
bivar_colors <- c(
  "1-1" = "#f8e8ef",
  "1-2" = "#e6b7ca",
  "1-3" = "#c77c9a",
  "2-1" = "#e0b5b9",
  "2-2" = "#c4788c",
  "2-3" = "#9e3f63",
  "3-1" = "#c78383",
  "3-2" = "#9b4655",
  "3-3" = "#7a003c"
)

#Note: Created bivariate choropleth map of PM2.5 and NO2 burden.
ggplot(bivar_map) +
  
  #Note: Added Chicago community polygons and colored them by pollution category.
  geom_sf(aes(fill = bivar_class), color = "gray35", linewidth = 0.25) +
  
  #Note: Applied custom bivariate color scale.
  scale_fill_manual(
    values = bivar_colors,
    name = "PM2.5 + NO2"
  ) +
  
  #Note: Added titles and labels.
  labs(
    title = "Combined PM2.5 and NO2 Burden",
    subtitle = "Darker areas represent communities with higher levels of both pollutants"
  ) +
  
  #Note: Applied minimal map styling.
  theme_void() +
  
  #Note: Customized title, subtitle, and legend appearance.
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 11
    ),
    legend.position = "right"
  )
```



## Row {data-height="120"}

### Mean PM2.5 rate
```{r}

#Note: Calculated the average PM2.5 concentration across all pollution sensors.
mean_pm25 <- round(
  mean(pollution$pm2_5ConcMass24HourMean.value, na.rm = TRUE),
  2
)

#Note: Created value box displaying the mean PM2.5 level.
valueBox(
  value = paste0(mean_pm25, " µg/m³"),
  caption = "Mean PM2.5 Level",
  icon = "fa-smog",
  color = "#7a003c"
)
```


### Mean No2 rate
```{r}

#Note: Calculated the average NO2 concentration across all pollution sensors.
mean_no2 <- round(
  mean(pollution$no2Conc24HourMean.value, na.rm = TRUE),
  2
)

#Note: Created value box displaying the mean NO2 level.
valueBox(
  value = paste0(mean_no2, " ppb"),
  caption = "Mean NO2 Level",
  icon = "fa-industry",
  color = "#006d2c"
)
```

## Row {data-height="450"}

### Overall Health and Pollution Hotspots across Chicago
```{r environmental-health-hotspots, echo=FALSE, warning=FALSE, message=FALSE, fig.width=11, fig.height=7}

#Note: Loaded libraries for data cleaning and spatial mapping.
library(tidyverse)
library(sf)

#Note: Stored GitHub links for pollution, health, and Chicago community boundary data.
pollution_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Open_Air_Chicago_Day_Aggregations_20260212(1)%20(3).csv"
health_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Chi_Health_Atlas_Data%20(1).csv"
community_url <- "https://data.cityofchicago.org/resource/igwz-8jzy.geojson"

#Note: Loaded pollution and health datasets.
pollution <- read_csv(pollution_url)
health <- read_csv(health_url)

#Note: Loaded Chicago community areas and standardized community names.
chicago_areas <- st_read(community_url, quiet = TRUE) %>%
  st_transform(4326) %>%
  mutate(
    area_numbe = as.numeric(area_numbe),
    community = str_to_title(community)
  )

#Note: Cleaned pollution sensor data and calculated average PM2.5 and NO2 values.
points <- pollution %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  group_by(sensor_name, latitude, longitude) %>%
  summarise(
    pm25 = mean(pm2_5ConcMass24HourMean.value, na.rm = TRUE),
    no2 = mean(no2Conc24HourMean.value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  #Note: Created a standardized pollution burden score from PM2.5 and NO2.
  mutate(
    pollution_burden =
      as.numeric(scale(pm25)) +
      as.numeric(scale(no2))
  ) %>%
  
  #Note: Converted pollution sensor locations into spatial points.
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

#Note: Matched pollution sensor points to Chicago community areas.
joined <- st_join(points, chicago_areas, join = st_within)

#Note: Calculated average pollution burden for each community area.
pollution_map <- joined %>%
  st_drop_geometry() %>%
  group_by(area_numbe, community) %>%
  summarise(
    pollution_burden = mean(pollution_burden, na.rm = TRUE),
    .groups = "drop"
  )

#Note: Created a health burden score using asthma, hypertension, and diabetes.
health_clean <- health %>%
  mutate(
    Name = str_to_title(Name),
    health_burden =
      as.numeric(scale(`HCSATH_2023-2024`)) +
      as.numeric(scale(`HCSHYT_2023-2024`)) +
      as.numeric(scale(`HCSDIA_2023-2024`))
  ) %>%
  select(Name, health_burden)

#Note: Combined pollution burden and health burden by community area.
combined_data <- pollution_map %>%
  left_join(
    health_clean,
    by = c("community" = "Name")
  ) %>%
  filter(
    !is.na(pollution_burden),
    !is.na(health_burden)
  ) %>%
  
  #Note: Created one overall burden score from pollution and health burden.
  mutate(
    combined_burden =
      as.numeric(scale(pollution_burden)) +
      as.numeric(scale(health_burden))
  )

#Note: Added Chicago community geometry back to the combined burden dataset.
combined_sf <- chicago_areas %>%
  left_join(combined_data, by = c("area_numbe", "community"))

#Note: Created a hexbin hotspot map using community centroid locations.
ggplot(
  combined_sf,
  aes(
    x = st_coordinates(st_centroid(geometry))[,1],
    y = st_coordinates(st_centroid(geometry))[,2],
    z = combined_burden
  )
) +
  
  #Note: Aggregated combined burden scores into hexagonal bins.
  stat_summary_hex(
    aes(fill = after_stat(value)),
    bins = 18,
    fun = mean,
    color = "white",
    linewidth = 0.25
  ) +
  
  #Note: Applied pink-white-green gradient for low, middle, and high burden.
  scale_fill_gradient2(
    low = "#f8c8dc",
    mid = "#f7f7f7",
    high = "#006d2c",
    midpoint = 0,
    name = "Combined\nBurden"
  ) +
  
  #Note: Added titles and axis labels.
  labs(
    title = "Overall burden combines PM2.5, NO2, asthma, hypertension, and diabetes",
    subtitle = "Pink = lower burden, green = higher combined burden across Chicago",
    x = "Longitude",
    y = "Latitude"
  ) +
  
  #Note: Applied a clean minimal theme.
  theme_minimal() +
  
  #Note: Customized title, subtitle, and grid appearance.
  theme(
    plot.title = element_text(
      face = "bold",
      size = 20,
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 11,
      hjust = 0.5
    ),
    panel.grid.minor = element_blank()
  )
```

### Intra-Categorical Correlation Plot: Disease
```{r pollution-health-quadrant, echo=FALSE, warning=FALSE, message=FALSE, fig.width=11, fig.height=7}

#Note: Loaded libraries for data cleaning and spatial mapping.
library(tidyverse)
library(sf)

#Note: Stored GitHub links for pollution, health, and Chicago community boundary data.
pollution_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Open_Air_Chicago_Day_Aggregations_20260212(1)%20(3).csv"
health_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Chi_Health_Atlas_Data%20(1).csv"
community_url <- "https://data.cityofchicago.org/resource/igwz-8jzy.geojson"

#Note: Loaded pollution and health datasets.
pollution <- read_csv(pollution_url)
health <- read_csv(health_url)

#Note: Loaded Chicago community areas and standardized community names.
chicago_areas <- st_read(community_url, quiet = TRUE) %>%
  st_transform(4326) %>%
  mutate(
    area_numbe = as.numeric(area_numbe),
    community = str_to_title(community)
  )

#Note: Cleaned pollution sensor data and calculated average PM2.5 and NO2 values.
pollution_points <- pollution %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  group_by(sensor_name, latitude, longitude) %>%
  summarise(
    pm25 = mean(pm2_5ConcMass24HourMean.value, na.rm = TRUE),
    no2 = mean(no2Conc24HourMean.value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  #Note: Removed sensor points missing PM2.5 or NO2 values.
  filter(!is.na(pm25), !is.na(no2)) %>%
  
  #Note: Created a standardized pollution burden score from PM2.5 and NO2.
  mutate(
    pollution_burden = as.numeric(scale(pm25)) + as.numeric(scale(no2))
  ) %>%
  
  #Note: Converted pollution sensor locations into spatial points.
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

#Note: Matched pollution sensor points to Chicago community areas.
pollution_area <- st_join(pollution_points, chicago_areas, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(area_numbe, community) %>%
  summarise(
    pollution_burden = mean(pollution_burden, na.rm = TRUE),
    .groups = "drop"
  )

#Note: Created a health burden score using asthma, hypertension, and diabetes.
health_clean <- health %>%
  mutate(
    community = str_to_title(Name),
    health_burden =
      as.numeric(scale(`HCSATH_2023-2024`)) +
      as.numeric(scale(`HCSHYT_2023-2024`)) +
      as.numeric(scale(`HCSDIA_2023-2024`))
  ) %>%
  select(community, health_burden)

#Note: Combined pollution and health burden data by community area.
quadrant_data <- pollution_area %>%
  left_join(health_clean, by = "community") %>%
  filter(!is.na(pollution_burden), !is.na(health_burden)) %>%
  
  #Note: Categorized communities into four burden groups using median cutoffs.
  mutate(
    category = case_when(
      pollution_burden >= median(pollution_burden, na.rm = TRUE) &
        health_burden >= median(health_burden, na.rm = TRUE) ~ "High pollution + high health burden",
      pollution_burden >= median(pollution_burden, na.rm = TRUE) &
        health_burden < median(health_burden, na.rm = TRUE) ~ "High pollution + lower health burden",
      pollution_burden < median(pollution_burden, na.rm = TRUE) &
        health_burden >= median(health_burden, na.rm = TRUE) ~ "Lower pollution + high health burden",
      TRUE ~ "Lower pollution + lower health burden"
    )
  )

#Note: Created quadrant plot comparing pollution burden and health burden.
ggplot(quadrant_data, aes(x = pollution_burden, y = health_burden)) +
  
  #Note: Added vertical median line to divide lower and higher pollution burden.
  geom_vline(
    xintercept = median(quadrant_data$pollution_burden, na.rm = TRUE),
    linetype = "dashed",
    color = "gray45"
  ) +
  
  #Note: Added horizontal median line to divide lower and higher health burden.
  geom_hline(
    yintercept = median(quadrant_data$health_burden, na.rm = TRUE),
    linetype = "dashed",
    color = "gray45"
  ) +
  
  #Note: Added points representing Chicago community areas.
  geom_point(
    aes(fill = category),
    shape = 21,
    size = 5,
    color = "white",
    alpha = 0.9
  ) +
  
  #Note: Applied custom colors for the four burden categories.
  scale_fill_manual(
    values = c(
      "High pollution + high health burden" = "#006d2c",
      "High pollution + lower health burden" = "#7a003c",
      "Lower pollution + high health burden" = "#f8c8dc",
      "Lower pollution + lower health burden" = "#d9f0d3"
    )
  ) +
  
  #Note: Added plot titles and axis labels.
  labs(
    title = "Where Pollution and Health Burdens Align",
    subtitle = "Each point represents a Chicago community area",
    x = "Pollution Burden Score",
    y = "Health Burden Score",
    fill = "Community Type"
  ) +
  
  #Note: Applied minimal theme styling.
  theme_minimal() +
  
  #Note: Customized title, subtitle, and legend placement.
  theme(
    plot.title = element_text(face = "bold", size = 19, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom"
  )
```



Obesity
=====================================

## Row {data-height="450"}

### Obesity Mismatch Index

``` {r obesity residual map and coef plot}
# Fit a linear regression model obesity ~ hardship
ob_model <- lm(obesity_rate ~ hard_index, data = chicago_map)
# extract model
ob_tbl <- tidy(ob_model, conf.int = TRUE)
# add residuals to df
chicago_map$residual[as.numeric(rownames(ob_model$model))] <- residuals(ob_model)
# Remove intercept and plot pointranges (estimate + CI) with a vertical line at 0
ob_tbl <- ob_tbl %>%
  filter(term != "(Intercept)")

# plot linear model
p_ob_hard <- ggplot(chicago_map, aes(x = obesity_rate, y = hard_index,
                                    color = residual,
                                    text = paste0("<b>", community, "</b><br>",
                                                  "Obesity Rate: ", round(obesity_rate, 1), "%<br>",
                                                  "Hardship: ", round(hard_index, 1), "<br>",
                                                  "Residual: ", round(residual, 1)))) +
  geom_point() +
  scale_color_gradient2(low = "#48bea1", mid = "#f4bb8f", high = "#7b4419",
                        midpoint = 0) +
  theme_minimal() +
  labs(title = "Obesity Prevalence Correlates with Socioeconomic Hardship",
       x = "Obesity Rate",
       y = "Hardship Index")

# interactive scatterplot
ggplotly(p_ob_hard, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)

# ggplot of regression coef plot
ob_hard <- ggplot(ob_tbl, aes(x = term, y = estimate)) +
  geom_pointrange(aes( ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "#e03000") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Coefficient plot with 95% CI",
       y = "Estimate",
       x = "Term")

# interactive coefficient plot 
ggplotly(ob_hard) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

### Obesity Mismatch Index

``` {r obesity residuals}
# residual tooltip
chicago_map <- chicago_map %>%
  filter(!is.na(obesity_rate)) %>%
  mutate(
    tooltip_text = paste0(
      "<b>", community, "</b><br>",
      "Obesity Rate: ", round(obesity_rate, 1), "%<br>",
      "Predicted: ", round(obesity_rate - residual, 1), "%<br>",
      "Residual: ", round(residual, 1)
    )
  )

ob_resid <- ggplot(chicago_map) +
  geom_sf(aes(fill = residual, text = tooltip_text)) +
  scale_fill_gradient2(
    low = "#48bea1", mid = "white", high = "#7b4419",
    midpoint = 0, name = "Residual"
  ) +
  theme_minimal() +
  labs(title = "Obesity Mismatch Index",
       subtitle = "Dark gray = worse than expected | Red = better than expected")

ggplotly(ob_resid, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Resilient Neighborhoods

```{r}
n_resilient <- sum(chicago_map$resilient_ob == TRUE, na.rm = TRUE)
valueBox(
  value = n_resilient,
  caption = "Resilient Neighborhoods",
  icon = "fa-shield",
  color = "#48bea1"
)
```

### Vulnerable Neighborhoods

```{r}
n_vulnerable <- sum(chicago_map$vulnerable_ob == TRUE, na.rm = TRUE)
valueBox(
  value = n_vulnerable,
  caption = "Vulnerable Neighborhoods",
  icon = "fa-triangle-exclamation",
  color = "#7b4419"
)
```

### Neighborhoods Outside Expected Pattern

```{r}
n_mismatch <- sum(chicago_map$resilient_ob == TRUE | chicago_map$vulnerable_ob == TRUE, na.rm = TRUE)
valueBox(
  value = n_mismatch,
  caption = "Neighborhoods Outside Expected Pattern",
  icon = "fa-shuffle",
  color = "#a7e831"
)
```

### Variance in Obesity Explained by Hardship

```{r}
valueBox(
  value = "39%",
  caption = "Variance in Obesity Explained by Hardship",
  icon = "fa-chart-line",
  color = "#167b2b"
)
```

## Row {data-height="450"}

### Neighborhoods of Vulnerability and Resilience

```{r mismatch scatterplot}
scatter_mismatch <- ggplot(chicago_map, aes(x = hard_index, y = disease_burden,
                                             color = category_ob,
                                             text = paste0("<b>", community, "</b><br>",
                                                           "Hardship: ", round(hard_index, 1), "<br>",
                                                           "Disease Burden: ", round(disease_burden, 1), "%<br>",
                                                           category_ob))) +
  geom_point(size = 3) +
  geom_vline(xintercept = median(chicago_map$hard_index, na.rm = TRUE), 
             linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = median(chicago_map$disease_burden, na.rm = TRUE), 
             linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Resilient" = "#48bea1", 
                                "Vulnerable" = "#D10300", 
                                "Expected" = "grey70")) +
  labs(x = "Hardship Index", y = "Disease Burden (%)", color = NULL) +
  theme_minimal()

ggplotly(scatter_mismatch, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

### Obesity-Specific Vulnerability and Resilience

``` {r scatterplot of diabetes vs hardship}
scatter_obesity <- ggplot(chicago_map, aes(x = hard_index, y = obesity_rate,
                                             color = category_ob,
                                             text = paste0("<b>", community, "</b><br>",
                                                           "Hardship: ", round(hard_index, 1), "<br>",
                                                           "Obesity Rate: ", round(obesity_rate, 1), "%<br>",
                                                           category_ob))) +
  geom_point(size = 3) +
  geom_vline(xintercept = median(chicago_map$hard_index, na.rm = TRUE), 
             linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = median(chicago_map$obesity_rate, na.rm = TRUE), 
             linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Resilient" = "#48bea1", 
                                "Vulnerable" = "#7b4419", 
                                "Expected" = "grey70")) +
  scale_x_discrete(limits = rev) +
  labs(x = "Hardship Index", y = "Obesity Rate (%)", color = NULL) +
  theme_minimal()

ggplotly(scatter_obesity, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

Obesity & Environment
=====================================

```{r obesity-env-setup, echo=FALSE, warning=FALSE, message=FALSE}

data_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Open_Air_Chicago_Day_Aggregations_20260212(1)%20(3).csv"

pollution_raw <- read_csv(data_url, show_col_types = FALSE)

area_air <- pollution_raw %>%
  mutate(community = str_to_title(str_trim(str_remove(sensor_name, "\\s*\\d+$")))) %>%
  filter(!str_detect(community, "Collocation")) %>%
  group_by(community) %>%
  summarise(
    no2_mean  = mean(no2Conc24HourMean.value, na.rm = TRUE),
    pm25_mean = mean(pm2_5ConcMass24HourMean.value, na.rm = TRUE),
    .groups   = "drop"
  )

chicago_env <- chicago_map %>%
  mutate(community_key = str_to_title(community)) %>%
  left_join(area_air, by = c("community_key" = "community")) %>%
  filter(!is.na(no2_mean), !is.na(obesity_rate),
         !is.na(hard_index), !is.na(pct_white))

baseline_mod <- lm(obesity_rate ~ pct_white + TRF_2020, data = chicago_env)
baseline_r2  <- round(summary(baseline_mod)$r.squared, 3)

chicago_env <- chicago_env %>%
  mutate(resid_baseline = residuals(baseline_mod))

full_mod  <- lm(obesity_rate ~ pct_white + TRF_2020 + no2_mean, data = chicago_env)
full_r2   <- round(summary(full_mod)$r.squared, 3)
no2_p     <- round(tidy(full_mod)$p.value[tidy(full_mod)$term == "no2_mean"], 3)
no2_coef  <- round(tidy(full_mod)$estimate[tidy(full_mod)$term == "no2_mean"], 3)

r_pw        <- round(cor(chicago_env$pct_white,    chicago_env$obesity_rate,   use = "complete.obs"), 3)
r_no2       <- round(cor(chicago_env$no2_mean,     chicago_env$obesity_rate,   use = "complete.obs"), 3)
r_no2_resid <- round(cor(chicago_env$no2_mean,     chicago_env$resid_baseline, use = "complete.obs"), 3)
```

## Row {data-height="500"}

### Introduction

**Background**

Obesity is among the most pressing public health challenges in the United States, having increased from 30.5% to 41.9% between 1999 and 2017 (CDC, 2022). Its etiology involves genetic predisposition, behavioral factors, and residential environment. Neighborhood-level determinants including poverty, unemployment, crime, housing instability, and environmental exposures have all been implicated in shaping the geographic distribution of obesity (Lotfata et al., 2022).

A growing body of evidence links environmental pollution to metabolic dysfunction. Ambient air pollutants, persistent organic pollutants (POPs), and heavy metals have been proposed as obesogenic agents operating through endocrine disruption, systemic inflammation, epigenetic modification, and dysregulation of adipogenesis (Khalil et al., 2023). PM2.5 and NO2 — both products of traffic-related combustion — have been associated with adipokine disruption, insulin resistance, and increased BMI in children and adults (Simkova et al., 2020).

Chicago exhibits pronounced geographic disparities in both pollution burden and chronic disease outcomes. South and west side neighborhoods experience substantially higher rates of obesity alongside elevated socioeconomic hardship (Lotfata et al., 2022). Whether ambient air quality contributes to obesity disparities beyond what socioeconomic variables explain has not yet been investigated at the community-area scale.

**Research Question:** To what extent does racial composition predict obesity disparities across Chicago community areas, and do neighborhoods with higher NO2 pollution burden carry excess obesity risk beyond what demographics alone explain?

**Hypothesis:** Neighborhoods with lower white population share will exhibit higher obesity rates. NO2 will show a modest positive association with obesity residuals after demographic controls.

### Methods

**Data Sources**

*Open Air Chicago* — Clarity Node-S sensor network deployed across all 77 Chicago community areas. Daily aggregated readings for PM2.5 and NO2 were used for September 2025 through February 2026 (n = 45,612 sensor-day observations). Community-area means were computed by averaging all valid daily readings per community area.

*Chicago Health Atlas (CHA)* — Community-area health and demographic estimates for all 77 community areas. Key variables: obesity prevalence (HCSOB 2023–2024), percent white population (PCT-W 2020–2024), hardship index (HDX 2020–2024), and traffic risk factor (TRF 2020).

*Chicago Community Area Shapefiles* — Polygon geometries from the City of Chicago Data Portal, joined by community area number.

All analyses conducted in R.

## Row {data-height="80"}

### Obesity–Race Correlation

```{r vb-r-pw, echo=FALSE}
valueBox(
  value   = r_pw,
  caption = "Pearson r: % White vs Obesity",
  icon    = "fa-chart-line",
  color   = "#7b4419"
)
```

### Baseline Model R²

```{r vb-r2-base, echo=FALSE}
valueBox(
  value   = paste0(round(baseline_r2 * 100, 1), "%"),
  caption = "Variance Explained by Demographics Alone",
  icon    = "fa-percent",
  color   = "#167b2b"
)
```

### NO2 vs Obesity (r)

```{r vb-r-no2, echo=FALSE}
valueBox(
  value   = r_no2,
  caption = "Pearson r: NO2 vs Obesity Rate",
  icon    = "fa-wind",
  color   = "#48bea1"
)
```

### NO2 Added to Model

```{r vb-r2-full, echo=FALSE}
valueBox(
  value   = paste0(round(full_r2 * 100, 1), "%"),
  caption = "Variance Explained After Adding NO2",
  icon    = "fa-plus",
  color   = "#7b4419"
)
```

### NO2 p-value

```{r vb-no2-p, echo=FALSE,}
valueBox(
  value   = no2_p,
  caption = "p-value: NO2 Coefficient",
  icon    = "fa-flask",
  color   = "#cccccc"
)
```

## Row {data-height="500"}

### Obesity Rate by Community Area

```{r map-obesity, echo=FALSE, warning=FALSE, message=FALSE}
ob_map <- ggplot(chicago_env) +
  geom_sf(
    aes(
      fill = obesity_rate,
      text = paste0(
        "<b>", community, "</b><br>",
        "Obesity rate: ", round(obesity_rate, 1), "%<br>",
        "% White: ", round(pct_white, 1), "%<br>",
        "NO2: ", round(no2_mean, 2), " ppb"
      )
    ),
    color = "gray40", linewidth = 0.2
  ) +
  scale_fill_gradient(low = "white", high = "#7b4419", name = "Obesity (%)") +
  labs(title = "Obesity Prevalence") +
  theme_void() +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 12, color = "#2F2F2F"),
    legend.position = "bottom"
  )

ggplotly(ob_map, tooltip = "text") %>%
  config(displayModeBar = FALSE) %>%
  layout(autosize = TRUE)
```

### % White Population

```{r map-pctwhite, echo=FALSE, warning=FALSE, message=FALSE}
pw_map <- ggplot(chicago_env) +
  geom_sf(
    aes(
      fill = pct_white,
      text = paste0(
        "<b>", community, "</b><br>",
        "% White: ", round(pct_white, 1), "%<br>",
        "Obesity rate: ", round(obesity_rate, 1), "%"
      )
    ),
    color = "gray40", linewidth = 0.2
  ) +
  scale_fill_gradient(low = "#d73027", high = "white", name = "% White") +
  labs(title = "Racial Composition (% White)") +
  theme_void() +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 12, color = "#2F2F2F"),
    legend.position = "bottom"
  )

ggplotly(pw_map, tooltip = "text") %>%
  config(displayModeBar = FALSE) %>%
  layout(autosize = TRUE)
```

### Mean NO2 by Community Area

```{r map-no2, echo=FALSE, warning=FALSE, message=FALSE}
no2_map <- ggplot(chicago_env) +
  geom_sf(
    aes(
      fill = no2_mean,
      text = paste0(
        "<b>", community, "</b><br>",
        "Mean NO2: ", round(no2_mean, 2), " ppb<br>",
        "Obesity rate: ", round(obesity_rate, 1), "%"
      )
    ),
    color = "gray40", linewidth = 0.2
  ) +
  scale_fill_gradient(low = "white", high = "#48bea1", name = "NO2 (ppb)") +
  labs(title = "Mean NO2 Concentration") +
  theme_void() +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 12, color = "#2F2F2F"),
    legend.position = "bottom"
  )

ggplotly(no2_map, tooltip = "text") %>%
  config(displayModeBar = FALSE) %>%
  layout(autosize = TRUE)
```

## Row {data-height="300"}

### Map Interpretation

The three maps reveal a striking geographic contrast. Obesity is heavily concentrated on Chicago's South and West Sides, mirroring the inverse distribution of white population share. This spatial overlap is the visual signature of the r = `r r_pw` correlation between racial composition and obesity. The NO2 map shows near-uniform concentrations across the city (range: 12.45–18.44 ppb, SD = 1.16 ppb) with no spatial clustering that mirrors the obesity pattern, already suggesting that sensor-derived NO2 at this scale will not explain obesity disparities.

## Row {data-height="500"}

### Race vs Obesity

```{r scatter-race, echo=FALSE, warning=FALSE, message=FALSE}
scatter_race <- ggplot(
  chicago_env %>% st_drop_geometry(),
  aes(x = pct_white, y = obesity_rate)
) +
  geom_smooth(method = "lm", se = TRUE, color = "#7b4419", fill = "#7b4419", alpha = 0.15) +
  geom_point(
    aes(
      color = obesity_rate,
      text  = paste0(
        "<b>", community, "</b><br>",
        "% White: ", round(pct_white, 1), "%<br>",
        "Obesity: ", round(obesity_rate, 1), "%"
      )
    ),
    size = 3, alpha = 0.85
  ) +
  scale_color_gradient(low = "#f4bb8f", high = "#7b4419", guide = "none") +
  labs(
    title = paste0("% White vs Obesity Rate (r = ", r_pw, ")"),
    x     = "% White Population",
    y     = "Obesity Rate (%)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12, color = "#2F2F2F"))

ggplotly(scatter_race, tooltip = "text") %>%
  config(displayModeBar = FALSE) %>%
  layout(autosize = TRUE)
```

### NO2 vs Obesity Residuals

```{r scatter-no2-resid, echo=FALSE, warning=FALSE, message=FALSE}
resid_scatter <- ggplot(
  chicago_env %>% st_drop_geometry(),
  aes(x = no2_mean, y = resid_baseline)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "lm", se = TRUE, color = "#2166ac", fill = "#2166ac", alpha = 0.15) +
  geom_point(
    aes(
      color = resid_baseline,
      text  = paste0(
        "<b>", community, "</b><br>",
        "NO2: ", round(no2_mean, 2), " ppb<br>",
        "Residual: ", round(resid_baseline, 2), "<br>",
        "Observed obesity: ", round(obesity_rate, 1), "%"
      )
    ),
    size = 3, alpha = 0.85
  ) +
  scale_color_gradient2(
    low = "#2166ac", mid = "gray80", high = "#d73027",
    midpoint = 0, guide = "none"
  ) +
  labs(
    title = paste0("NO2 vs Obesity Residuals (r = ", r_no2_resid, ", p = ", no2_p, ")"),
    x     = "Mean NO2 (ppb)",
    y     = "Residual (observed − predicted obesity)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12, color = "#2F2F2F"))

ggplotly(resid_scatter, tooltip = "text") %>%
  config(displayModeBar = FALSE) %>%
  layout(autosize = TRUE)
```

## Row {data-height="300"}

### Scatter Interpretation

**Left:** Percent white population is a strong negative predictor of obesity (r = `r r_pw`, p < 0.001). The relationship is linear and consistent, with predominantly white neighborhoods clustered at low obesity rates and predominantly non-white neighborhoods at high rates. This single variable accounts for `r round(baseline_r2*100,1)`% of community-area obesity variance when combined with traffic risk. **Right:** After removing demographic variance, NO2 shows no meaningful relationship with residual obesity disparity (r = `r r_no2_resid`, p = `r no2_p`). The flat regression line confirms that measured NO2 at this spatiotemporal resolution does not predict excess obesity burden.

## Row {data-height="380"}

### Regression Coefficient Table

```{r coef-table, echo=FALSE, warning=FALSE, message=FALSE}
full_tidy <- tidy(full_mod, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    Term      = case_when(
      term == "pct_white" ~ "% White Population",
      term == "TRF_2020"  ~ "Traffic Risk Factor",
      term == "no2_mean"  ~ "Mean NO2 (ppb)",
      TRUE                ~ term
    ),
    Estimate  = round(estimate, 3),
    `95% CI`  = paste0("[", round(conf.low, 3), ", ", round(conf.high, 3), "]"),
    `p-value` = round(p.value, 3),
    Sig.      = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    )
  ) %>%
  select(Term, Estimate, `95% CI`, `p-value`, Sig.)

DT::datatable(
  full_tidy,
  rownames = FALSE,
  options  = list(dom = "t", ordering = FALSE, pageLength = 5),
  caption  = htmltools::tags$caption(
    style = "caption-side:top; font-weight:bold; font-size:0.9rem;",
    "Full Model: Obesity ~ % White + Traffic Risk + NO2"
  )
) %>%
  DT::formatStyle(
    "Sig.",
    color = DT::styleEqual(
      c("***", "**", "*", "ns"),
      c("#d73027", "#d73027", "#f4a442", "#aaaaaa")
    )
  )
```

### Discussion

**Key Findings**

The dominant finding of this analysis is that racial composition is the strongest predictor of obesity disparities across Chicago community areas (r = `r r_pw`). Low % White neighborhoods such as those concentrated on the south and west sides consistently exhibit higher obesity rates, possibly reflecting the broader effects of structural disinvestment, neighborhood poverty, limited food access, and chronic stress (Lotfata et al., 2022). Percent white and traffic risk together explain approximately `r round(baseline_r2*100,1)`% of community-area obesity variance, consistent with previous ecological analyses of Chicago.

Contrary to our hypothesis, **NO2 does not predict obesity burden after controlling for demographics (β = `r no2_coef`, p = `r no2_p`).** Adding NO2 increases model R² by less than 0.3 percentage points. This null result should not be interpreted as evidence that air pollution is irrelevant to obesity. Key constraints include: the observation window covers only one fall/winter season; community-area aggregation erases intra-neighborhood variation; and NO2 spatial variance across the network is extremely low (SD = 1.16 ppb), limiting statistical power regardless of true effect magnitude.

**Limitations**

Chicago Air data provides only a six-month window, a time period insufficient for predicting the effects of pollution on longterm outcomes for chronic diseases such as obesity. Individual-level data would reduce ecological confounding. Variables absent from this dataset such as food environment, green space, crime, housing instability, and endocrine-disrupting pollutants such as PAHs and POPs — likely contribute to the ~46.7%% of variance not explained by the model (Khalil et al., 2023; Simkova et al., 2020).

**References:**
Khalil, W. J., Akeblersane, M., Khan, A. S., Moin, A. S. M., & Butler, A. E. (2023). Environmental pollution and the risk of developing metabolic disorders: Obesity and diabetes. International Journal of Molecular Sciences, 24(10), 8870. https://doi.org/10.3390/ijms24108870

Lotfata, A., Georganos, S., Kalogirou, S., & Helbich, M. (2022). Ecological associations between obesity prevalence and neighborhood determinants using spatial machine learning in Chicago, Illinois, USA. ISPRS International Journal of Geo-Information, 11(11), 550. https://doi.org/10.3390/ijgi11110550

Simkova, S., Veleminsky, M., & Sram, R. J. (2020). The impact of air pollution to obesity. Neuroendocrinology Letters, 41(3), 146–153.

Hypertension
=====================================

Hypertension is used as a central indicator because it reflects long-term exposure to both environmental and structural conditions, making it a powerful measure of inequality across space.

## Row {data-height="600"}

```{r load data, include=FALSE}

### Data Preparation (hypertension)

#Note: These data sets come from the Chicago Health Atlas

# load data
url3 <- "https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/Chi_Health_Atlas_Data(1).csv"
health_data <- read.csv(url3)
##### health_data <- read_csv("Chi_Health_Atlas_Data(1).csv")

url4 <- "https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/Chicago%20Health%20Atlas%20Data%20Download%20-%20Census%20Tracts%20(2).csv"
poverty_data <- read.csv(url4)

##### poverty_data <- read_csv("Chicago Health Atlas Data Download - Census Tracts (2).csv")
```

```{r clean-data, include=FALSE}
#Note: I selected only the variables needed for analysis.
#Note: This reduces errors and keeps the data set focused.

analysis_df <- health_data %>%
  select(
    community = Name,
    lat = Latitude,
    lon = Longitude,
    hypertension = `HCSHYT_2023.2024`,
    pm25 = `PMC_2020`,
    population = `POP_2020.2024`
  ) %>%
  mutate(community = toupper(str_trim(community)))

#Note: Cleaned poverty data for merging
poverty_data <- poverty_data %>%
  select(
    community = Name,
    poverty_rate = `POV_2020.2024`
  ) %>%
  mutate(community = toupper(str_trim(community)))

#Note: Joined data sets
analysis_df <- analysis_df %>%
  left_join(poverty_data, by = "community")

#Note: Ensured poverty is numeric (prevents plotting errors)
analysis_df$poverty_rate <- readr::parse_number(analysis_df$poverty_rate)
```

### Environmental Exposure (hypertension)

ANALYSIS

PM2.5 exposure across Chicago demonstrates clear clusters of elevated pollution levels across specific neighborhoods. These areas represent communities that are consistently subjected to higher environmental risk, which can contribute to long-term health consequences. The non-random distribution of pollution suggests that environmental burden is structurally embedded within the urban landscape. This pattern raises important concerns about environmental justice and unequal exposure. On average, a PM2.5 of around 9.2 suggests moderate air quality across neighborhoods in Chicago; however a slight health risk is present for individuals sensitive to pollutants and prone to having chronic diseases. Establishing this baseline is critical for interpreting how environmental conditions shape health outcomes.

### PM2.5 Map: Environmental Risk Baseline

```{r pm25_map}
#Note: This PM2.5 map establishes environmental risk baseline

pm_df <- analysis_df %>%
  filter(!is.na(lat), !is.na(lon), !is.na(pm25))

plot_ly(
  pm_df,
  type = "scattermapbox",
  lat = ~lat,
  lon = ~lon,
  mode = "markers",
  marker = list(
    size = 10,
    color = ~pm25,
    colorscale = list(
      c(0, sun_yellow),
      c(0.5, sun_orange),
      c(1, sun_red)
    ),
    showscale = TRUE,
    line = list(color = dark_gray, width = 1)
  ),
  text = ~paste("Community:", community,
                "<br>PM2.5:", round(pm25,2))
) %>%
layout(
  mapbox = list(
    style = "carto-positron",
    zoom = 9,
    center = list(lat = 41.85, lon = -87.68)
  )
)
```


## Row {data-height="800"}

### Hypertension Analysis (spatial distribution) (hypertension)

ANALYSIS

Hypertension exhibits strong spatial clustering across Chicago, with certain neighborhoods consistently experiencing higher rates than others such as the community in Austin with 27,500 cases recorded compared to Riverdale with 2000 recorded cases. This pattern suggests that health outcomes are shaped by localized structural conditions rather than random variation. The persistence of these clusters indicate long-term exposure to risk factors such as economic stress and limited access to healthcare. Not all high-risk areas align perfectly with pollution patterns, pointing to additional underlying influences. This reinforces the importance of area-based analysis in understanding health disparities.

### Spatial Distribution of Hypertension

```{r mapbox-hypertension}
#Note:It uses Plotly's Mapbox engine for smooth zooming and better aesthetics.

#Note: IMPORTANT:
#Note: filtered only necessary variables to avoid NA issues
map_df <- analysis_df %>%
  filter(!is.na(lat), !is.na(lon), !is.na(hypertension))

#Note: Create interactive map
plot_ly(
  data = map_df,
  type = "scattermapbox",
  lat = ~lat,
  lon = ~lon,
  mode = "markers",

  #Note: Color encodes hypertension intensity
  marker = list(
    size = 10,
    color = ~hypertension,
    colorscale = list(
      c(0, "#FFC300"),   # yellow
      c(0.5, "#FF7A00"), # orange
      c(1, "#FF3B3B")    # red
    ),
    showscale = TRUE,
    line = list(color = "#2F2F2F", width = 1)
  ),

  #Note: Tooltip info
  text = ~paste(
    "Community:", community,
    "<br>Hypertension:", round(hypertension, 2)
  ),
  hoverinfo = "text"
) %>%
layout(
  mapbox = list(
    style = "carto-positron",   #Note: clean grayscale base map
    zoom = 9,
    center = list(lat = 41.85, lon = -87.68)
  ),
  margin = list(l = 0, r = 0, t = 40, b = 0),
  title = "Spatial Distribution of Hypertension Across Chicago"
)
```


## Row

### Pollution Relationship (hypertension)

ANALYSIS

The relationship between PM2.5 and hypertension shows a general downward trend when observing PM2.5 of 8 to 9.8, indicating that environmental exposure does slightly contribute to health risk. However, the variability around the trend line suggests that this relationship is not significant and realistic. Some neighborhoods experience higher-than-expected hypertension despite lower pollution levels. This indicates that additional structural or social factors are influencing outcomes. The results highlight the limitations of relying solely on environmental variables to explain health disparities.

```{r pollution-scatter}
#Note: This tests the expected environmental relationship.

scatter_df <- analysis_df %>%
  filter(!is.na(pm25), !is.na(hypertension))

p <- ggplot(scatter_df, aes(pm25, hypertension)) +
  geom_point(color = sun_orange, alpha = 0.7) +
  geom_smooth(method = "lm", color = sun_red) +
  theme_minimal()

ggplotly(p)
```


## Row {data-height="450"}

### Mismatch Data (hypertension)

ANALYSIS

The mismatch index highlights where observed hypertension diverges from expected patterns based on environmental and demographic factors. These deviations are spatially clustered, indicating localized influences that are beyond pollution. Areas with high mismatch values may experience structural disadvantages that can amplify and be prone to health risk. Conversely, lower-than-expected values suggest the presence of protective community factors across the city of Chicago. This approach can help provide a deeper understanding of inequality by identifying where standard explanations fall short.

```{r mismatch}
#Note: Residuals capture deviation from expected outcomes.

model <- lm(hypertension ~ pm25 + population, data = analysis_df)

analysis_df <- analysis_df %>%
  mutate(mismatch_index = resid(model))

```

### Mismatch Map

```{r mismatch-map}
#Note: plot is not that detailed. not how I would like to showcase the data but came across issues and decided to leave this plot as final product.

mismatch_df <- analysis_df %>%
  filter(!is.na(mismatch_index), !is.na(lat), !is.na(lon))

plot_ly(
  mismatch_df,
  x = ~lon,
  y = ~lat,
  type = "scatter",
  mode = "markers",
  color = ~mismatch_index,
  colors = c(sun_pink, sun_orange, sun_red)
)
```

## Row

### Conclusion (hypertension)
Hypertension across Chicago is shaped by both environmental exposure and structural inequality. While pollution contributes to risk, it does not fully explain the observed variation.

The mismatch framework reveals that health outcomes are influenced by a broader set of factors, including socioeconomic conditions and neighborhood context. These findings emphasize the need for comprehensive approaches to public health. 


### Future Directions (hypertension)

Future work can expand this analysis by incorporating additional variables such as access to healthcare, green space, and community-level trust.

Temporal analysis could reveal how these relationships evolve over time. More advanced spatial models may also better capture neighborhood-level effects.

Understanding these dynamics more deeply can help design targeted interventions that address both environmental and structural drivers of health inequality.


Asthma
=====================================
## Row {data-height="300"}
### Abstract (Asthma)

**Introduction: This research aims to explore weather traffic pollution disproportionately affects some Chicago neighborhoods more than others. According to the American Lung Association previous findings have revealed there are correlations between air quality/pollution and health outcomes in the city of Chicago, predominantly affecting densely populated, disadvantaged areas. Based on this research, we hypothesize neighborhoods with higher levels of environmental burden will be more prone to higher asthma levels.**

**Methods: Descriptive analysis was used in order to conduct data exploration research on Chicago pollution metrics and their correlation to health outcomes. Data from the Chicago Health Atlas and Chicago Housing Authority was used in order to create four figures using R Studio, packages include: ggplot, tidyverse, janitor, and plotly. Figures used included a leaflet map, mismatch graph, standard plot and environmental burden plot. Graph interpretation was used in order to conduct results.**

```{r}
#Note: Loaded data.
url5 <- "https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/Chi_Health_Atlas_Data.csv"
health_data <- read.csv(url5) %>%
  clean_names()

####health_data <- read_csv("Chi_Health_Atlas_Data.csv") %>%
####  clean_names()

#Note: Filtered through the only columns we needed."Name" includes the name of each neighborhood, "HCSATH_2023-2024" showcases the asthma and burden outcomes for each neighborhood, this data set will be used to showcase real health outcomes. "HCSATH_2023-2024" showcases the environmental burden caused by traffic which is used as a predictor of pollution to show expected health outcomes.
asthma_plot <- health_data %>%
  select(name,
    asthma = hcsath_2023_2024,
    pollution = trf_2020
  )
```

## Row {data-height="600"}

### Figure 1: Leaflet Map of Asthma Burden of Chicago (asthma)

**The leaflet map sets the baseline data that shows where asthma is distributed in the city. This map is interactive and represents the prevalence of asthma in 72 Chicago neighborhoods. Asthma concentration is represented in large yellow circles which decrease in size and gradience from dark green (greater concentrations) to light green (lower concentrations). Viewers can zoom into specific neighborhoods and hover over them to reveal neighborhood name, asthma prevalence and traffic pollution levels. This map helps us identify spatial distribution of asthma throughout the city. It also helps us identify weather there is a correlation among pollution and asthma for specific neighborhoods. The map shows us asthma levels have a relatively even distribution in the city of Chicago with some higher concentrations on the northside such as Lakeview and Albany Park. However, there is no clear relationship between traffic pollution and asthma prevalence as these factors vary throughout neighborhoods.** 

### Spatial Trends of Asthma in Chicago

```{r}
#### redundant so commented
# #Note: Loaded data and cleaned it.
# health_data <- read_csv("Chi_Health_Atlas_Data.csv") %>%
#   clean_names()

#Note: Included the variables needed for the map.
map_data <- health_data %>%
  select(
    name,
    latitude,
    longitude,
    asthma = hcsath_2023_2024,
    pollution = trf_2020
  ) %>%
  drop_na()

#Note: Created a green palette with various hues based on asthma prevalence, including my color choice for the project. 
pal <- colorNumeric(
  palette = c("#e8f5e9", "#a5d6a7", "#43a047", "#1b5e20"),
  domain = map_data$asthma
)

#Note: Built interactive asthma distribution plot using the map of Illinois and surrounding areas, making 72 Chicago neighborhoods interactive.
leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  #Note: Added grey background to emphasize neighborhoods, added neighborhood markers as well as longitude and latitude to map points to their correct location.
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    #Note: Used scale and gradience to ensure neighborhoods with larger/darker concentrations of of asthma would have larger interactive circles.
    radius = ~rescale(asthma, to = c(6, 18)),
    fillColor = ~pal(asthma),
    #Note: Customized the circles color, weight and stroke.
    fillOpacity = 0.9,
    color = "black",
    weight = 1,
    stroke = TRUE,
    #Note: Groups markers/circles together when zooming out and separate when zooming in. The legend also matches the color palette from above and matches data set.
     clusterOptions = markerClusterOptions(),
    popup = ~paste0(
      "<b>", name, "</b><br>",
      "Asthma prevalence: ", round(asthma, 2), "<br>",
      "Traffic pollution: ", round(pollution, 2)
    )
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~asthma,
    title = "Asthma prevalence",
    opacity = 0.9
  )
```
## Row {data-height="1000"}

### Figure 2: Mismatch Graph of Pollution Risk (asthma)

**Description: This graph was created in order to compare the expected health outcomes for 77 Chicago neighborhoods based on the environmental pollution outcomes to real asthma levels. Expected Asthma levels were showcased using the traffic burden data set on top where lime green represents positive values (>0) that matched expected outcomes. The bottom showcases resilient neighborhoods that did not match expected outcomes in the negative value range (<0) in dark green. Grey circles represent exactly where outcomes match expectations. The tooltip function helps the viewer see data on individual plot points. This is helpful due to our large data set. By hovering over each point, the viewer is able to directly verify the mismatch and confirm information.**

**Analysis :The graph suggests health outcomes for asthma are have a low-moderate correlation to traffic related pollution as there is a general positive trend with a few positive outliers such as Lakeview and Austin. However, there is variability in the dataset which could mean multiple factors contribute to asthma outcomes.**

### Pollution Risk

```{r, fig.height=20, fig.width=12}
#Note: loaded the data into R.
health_data <- read.csv(url5)

# Note: Filtered the data set to only the necessary columns, I renamed the asthma and pollution variables to create simpler names and I removed neighborhoods missing asthma or pollution data using "drop_na".
asthma_plot <- health_data %>%
  select(
    Name,
    asthma = HCSATH_2023.2024,
    pollution = TRF_2020
  ) %>%
  drop_na()

# Note: Created a linear model that predicts asthma from pollution based on the traffic burden, predicting what asthma should look like. Created a custom string for each plot point in order to incorporate interactivity, used tooltip in order to display data when hovering.
mismatch_model <- lm(asthma ~ pollution, data = asthma_plot)

asthma_plot <- asthma_plot %>%
  mutate(
    expected_asthma = predict(mismatch_model),
    mismatch = asthma - expected_asthma,
    category = case_when(
      mismatch > 0 ~ "More vulnerable than expected",
      mismatch < 0 ~ "More resilient than expected",
      TRUE ~ "About as expected"
    )
  ) %>%
  arrange(mismatch) %>%
  mutate(
    Name = factor(Name, levels = Name),
    hover_text = paste(
      "Neighborhood:", Name,
      "<br>Asthma:", round(asthma, 2),
      "<br>Expected asthma:", round(expected_asthma, 2),
      "<br>Mismatch:", round(mismatch, 2),
      "<br>Traffic burden:", round(pollution, 2),
      "<br>Category:", category
    )
  )

#Note: Created the graph, making x=mismatch value and y=neighborhood name, the fill showcases the resilience/vulnerability category.
p <- ggplot(asthma_plot, aes(x = mismatch, y = Name, fill = category, text = hover_text)) +
  
  # Note: Created horizontal bars for each neighborhood and made thinner bars in order to reduce crowding.
  geom_col(width = 0.5) +
  
  # Note: Added a vertical dashed line at 0 to separate higher than expected from lower than expected.
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.7, color = "black") +
  
  # Note: Included points based on pollution burden to see the amount of traffic burden in each neighborhood.
  geom_point(aes(size = pollution, text = hover_text), color = "black", alpha = 0.65) +
  
  # Note: Added white space to prevent crowding on the x-axis bars.
  scale_x_continuous(expand = expansion(mult = c(0.08, 0.12))) +
  
  # Note: Chose my custom color theme to represent each category.
  scale_fill_manual(values = c(
    "More vulnerable than expected" = "#92F96A",
    "More resilient than expected" = "#74AC64",
    "About as expected" = "gray70"
  )) +
  
  # Note: Added titles to each axis, legends, main title and subtitle.
  labs(
    title = "Asthma Burden Beyond Expected Pollution Risk in Chicago",
    subtitle = "Positive values show neighborhoods with higher asthma than predicted from traffic-related pollution burden",
    x = "Asthma mismatch (actual - expected)",
    y = NULL,
    fill = NULL,
    size = "Traffic burden"
  ) +
  
  # Note: Used a minimal theme and customized multiple elements in order to improve readability and avoid crowding.
geom_col(width = 0.4) +

scale_y_discrete(expand = expansion(mult = c(0.02, 0.02))) +

theme_minimal(base_size = 12) +
theme(
  panel.grid.major.y = element_blank(),
  panel.grid.minor = element_blank(),
  plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
  plot.subtitle = element_text(size = 11, hjust = 0.5),
  axis.text.y = element_text(size = 11),
  axis.text.x = element_text(size = 11),
  legend.title = element_text(size = 11),
  legend.text = element_text(size = 10),
  plot.margin = margin(15, 20, 15, 15)
)
#Note: Converted static ggplot into an interactive plot by including the tooltip feature.
ggplotly(p, tooltip = "text")

```


## Row

### Figure 3: Standard Scatter Plot of Asthama vs Traffic Pollution (asthma)

**Description: The graph demonstrates the relationship between traffic-related pollution and prevalence of asthma across 77 Chicago neighborhoods. The x-axis represents traffic-related pollution and the y-axis represents the percentage of adults with asthma (prevalence). The color gradient increases in hue to a darker green in correlation to higher pollution levels. The dashed lines represent city averages which divide the plot into four quadrants. Each plot point represents a specific Chicago neighborhood and a tooltip feature was used in order to showcase interactivity. Each plot point tells the viewer what neighborhood they are looking at, as well as the asthma and pollution levels.**

**Analysis: Overall, there is a general positive relationship in the data as plot points are scattered at an upwards direction, suggesting pollution could contribute to asthma levels. However, plot-points do not emphasize a linear relationship and are scattered, meaning there could be other contributing factors to asthma levels. Furthermore, the quadrants on the graph help reveal patterns, the top right quadrant show expected burden. The top left, unexpected vulnerability, bottom right: resilience and bottom left: highest resilience. Some outliers are observed but there is no clear pattern. In conclusion, the graph helps the viewer see vulnerability patterns beyond environmental exposure.**

### Scatter plot

```{r}
#Note: Loaded the data set and cleaned column names using the janitor package. 
health_data <- health_data %>%
  clean_names()

#Note: Selected only the variables needed for the graph. 
plot_data <- health_data %>%
  select(
    name,
    asthma = hcsath_2023_2024,
    pollution = trf_2020
  ) %>%
  drop_na()

#Note: Calculated averages for asthma and pollution across the city.
avg_asthma <- mean(plot_data$asthma, na.rm = TRUE)
avg_pollution <- mean(plot_data$pollution, na.rm = TRUE)

# Note: Created the base ggplot and added the x and y axes.
p <- ggplot(
  plot_data,
  aes(
    x = pollution,
    y = asthma,
    color = pollution,
    text = paste(
      "Neighborhood:", name,
      "<br>Pollution:", round(pollution, 2),
      "<br>Asthma:", round(asthma, 2)
    )
  )
) +
  
  #Note: Plotted each neighborhood as a seperate data point. 
  geom_point(size = 3, alpha = 0.85) +
  
  #Note: Added dashed reference lines at city averages
  geom_vline(xintercept = avg_pollution, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = avg_asthma, linetype = "dashed", color = "gray40") +
  
  #Note: Used my custom green gradient colors instead of category colors
  scale_color_gradientn(
    colors = c("#dff3e3", "#9bd49f", "#4caf50", "#1b5e20"),
    name = "Traffic Pollution"
  ) +
  
  #Note: Added axes titles, main title and labels, kept minimal theme.
  labs(
    title = "Asthma vs Traffic Pollution in Chicago Neighborhoods",
    x = "Traffic Pollution",
    y = "Asthma Prevalence"
  ) +
  

  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.0),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

#Note: Used a tooltip to make plot interactive.
ggplotly(p, tooltip = "text")
```


## Row

### Figure 4: Environmental Justice Burden Plot (asthma)

**Description: The graph showcases the relationship between the environmental justice burden of asthma and asthma mismatch rates. The graph plots environmental justice burden and asthma on the x-axis and asthma mismatch on the y-axis. Asthma mismatch is determined by actual minus expected asthma levels based on traffic pollution. The graph aims to explore weather certain neighborhoods are vulnerable and disproportionately affected by asthma. Neighborhoods over the dotted line (0>) have a higher than expected asthma level despite traffic pollution. Neighborhoods under the dotted line (<0) more resilient than expected asthma rates. The graph also includes an interactive feature that allows the viewer to hover over a specific neighborhood to see neighborhood name, environmental justice burden, mismatch score, asthma prevalence and pollution levels.** 

**Analysis: This graph suggests a slight positive relationship between higher levels of EJ burden and positive mismatch yields. However, variability still exists in the graph and distribution appears mostly even. It is possible, neighborhood plays a role into asthma levels.**

### Environmental Justice Burden and Asthma

```{r}
#Note: Loaded and cleaned data set.
#### health_data <- read_csv("Chi_Health_Atlas_Data.csv") %>%
####  clean_names()

#Note: Selected only the variables required for the graph. 
plot_data <- health_data %>%
  select(
    name,
    asthma = hcsath_2023_2024,
    pollution = trf_2020,
    ej_burden = chaknkc_2023
  ) %>%
  drop_na()

#Note: Incorporated mismatch model to calculate expected asthma levels based on traffic pollution.
mismatch_model <- lm(asthma ~ pollution, data = plot_data)

#Note: Calculated mismatch and categorized neighborhoods.
plot_data <- plot_data %>%
  mutate(
    expected_asthma = predict(mismatch_model),
    mismatch = asthma - expected_asthma,
    category = case_when(
      mismatch > 0 ~ "More vulnerable than expected",
      mismatch < 0 ~ "More resilient than expected",
      TRUE ~ "About as expected"
    ),
    hover_text = paste(
      "Neighborhood:", name,
      "<br>EJ burden:", ej_burden,
      "<br>Mismatch:", round(mismatch, 2),
      "<br>Asthma:", round(asthma, 2),
      "<br>Pollution:", round(pollution, 2),
      "<br>Category:", category
    )
  )

#Note: Created the mismatch plot.
p <- ggplot(plot_data,
            aes(x = ej_burden, y = mismatch,
                color = category,
                text = hover_text)) +
  
  #Note: Plotted points for each neighborhood.
  geom_point(size = 3, alpha = 0.8) +
  
  #Note: Added regression line to show overall trend
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
  
  #Note: Added the dashed horizontal line at zero to create the mismatch.
  geom_hline(yintercept = 0, linetype = "dashed") +
  
  #Note:Used custom green color scheme.
  scale_color_manual(values = c(
    "More vulnerable than expected" = "#2e7d32",
    "More resilient than expected" = "#81c784",
    "About as expected" = "gray70"
  )) +
  
  #Note: Added axis labels, a title, legends and subtitles, kept minimal theme.
  labs(
    title = "Environmental Justice Burden and Asthma Mismatch",
    subtitle = "Exploring whether vulnerable neighborhoods face disproportionate health impacts",
    x = "Environmental Justice Burden",
    y = "Asthma Mismatch (Actual - Expected)",
    color = "Category"
  ) +
  
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom"
  )

#Note: Converted to interactive by incorporating a tooltip.
ggplotly(p, tooltip = "text")
```


## Row {data-height="650"}

### Figure 5: Correlation Between Asthma and Pollution (No2 and PM2.5) Burden 
```{r asthma-pollution-correlation, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}

#Note: Loaded libraries for data cleaning, spatial mapping, and interactivity.
library(tidyverse)
library(sf)
library(plotly)

#Note: Stored GitHub links for pollution, health, and Chicago community boundary data.
pollution_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Open_Air_Chicago_Day_Aggregations_20260212(1)%20(3).csv"
health_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Chi_Health_Atlas_Data%20(1).csv"
community_url <- "https://data.cityofchicago.org/resource/igwz-8jzy.geojson"

#Note: Loaded pollution and health datasets.
pollution <- read_csv(pollution_url)
health <- read_csv(health_url)

#Note: Loaded Chicago community areas and standardized community names.
chicago_areas <- st_read(community_url, quiet = TRUE) %>%
  st_transform(4326) %>%
  mutate(
    area_numbe = as.numeric(area_numbe),
    community = str_to_title(community)
  )

#Note: Cleaned pollution sensor data and calculated average PM2.5 and NO2 values.
pollution_points <- pollution %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  group_by(sensor_name, latitude, longitude) %>%
  summarise(
    pm25 = mean(pm2_5ConcMass24HourMean.value, na.rm = TRUE),
    no2 = mean(no2Conc24HourMean.value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  #Note: Removed sensor points with missing PM2.5 or NO2 values.
  filter(!is.na(pm25), !is.na(no2)) %>%
  
  #Note: Created a standardized pollution burden score from PM2.5 and NO2.
  mutate(
    pollution_burden = as.numeric(scale(pm25)) + as.numeric(scale(no2))
  ) %>%
  
  #Note: Converted pollution sensor locations into spatial points.
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

#Note: Matched pollution sensor points to Chicago community areas.
pollution_area <- st_join(pollution_points, chicago_areas, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(area_numbe, community) %>%
  summarise(
    pollution_burden = mean(pollution_burden, na.rm = TRUE),
    .groups = "drop"
  )

#Note: Cleaned health data and selected asthma rate by community area.
health_clean <- health %>%
  mutate(
    community = str_to_title(Name),
    asthma_rate = `HCSATH_2023-2024`
  ) %>%
  select(community, asthma_rate)

#Note: Combined pollution burden and asthma rate into one dataset.
asthma_pollution <- pollution_area %>%
  left_join(health_clean, by = "community") %>%
  filter(!is.na(pollution_burden), !is.na(asthma_rate)) %>%
  
  #Note: Created hover text for the interactive plot.
  mutate(
    hover_text = paste0(
      "<b>", community, "</b>",
      "<br>Pollution Burden: ", round(pollution_burden, 2),
      "<br>Asthma Rate: ", round(asthma_rate, 2)
    )
  )

#Note: Calculated correlation between pollution burden and asthma rate.
cor_value <- round(
  cor(asthma_pollution$pollution_burden, asthma_pollution$asthma_rate, use = "complete.obs"),
  2
)

#Note: Created scatterplot showing asthma rate and pollution burden.
p <- ggplot(
  asthma_pollution,
  aes(
    x = pollution_burden,
    y = asthma_rate,
    text = hover_text
  )
) +
  
  #Note: Added community points colored by pollution burden.
  geom_point(aes(color = pollution_burden), size = 4.5, alpha = 0.85) +
  
  #Note: Added linear trend line to show the overall relationship.
  geom_smooth(method = "lm", se = TRUE, color = "#006d2c") +
  
  #Note: Applied pink-to-purple color gradient for pollution burden.
  scale_color_gradient(low = "#f8c8dc", high = "#7a003c") +
  
  #Note: Added title, subtitle, axes, and legend labels.
  labs(
    title = "Correlation Between Asthma and Pollution Burden",
    subtitle = paste0("Each point represents a Chicago community area | r = ", cor_value),
    x = "Pollution Burden Score",
    y = "Asthma Rate",
    color = "Pollution\nBurden"
  ) +
  
  #Note: Applied a clean minimal theme.
  theme_minimal()

#Note: Converted scatterplot into an interactive plotly visualization.
ggplotly(p, tooltip = "text")
``` 



### Figure 6: Correlation Between Asthma and Pollution (No2 and PM2.5) Burden, Leaflet Plot

```{r leaflet-pollution-asthma, echo=FALSE, warning=FALSE, message=FALSE}

#Note: Loaded libraries for data cleaning, spatial mapping, and interactive maps.
library(tidyverse)
library(sf)
library(leaflet)

#Note: Stored GitHub links for pollution, health, and Chicago community boundary data.
pollution_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Open_Air_Chicago_Day_Aggregations_20260212(1)%20(3).csv"
health_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Chi_Health_Atlas_Data%20(1).csv"
community_url <- "https://data.cityofchicago.org/resource/igwz-8jzy.geojson"

#Note: Loaded pollution and health datasets.
pollution <- read_csv(pollution_url)
health <- read_csv(health_url)

#Note: Loaded Chicago community areas and standardized community names.
chicago_areas <- st_read(community_url, quiet = TRUE) %>%
  st_transform(4326) %>%
  mutate(
    area_numbe = as.numeric(area_numbe),
    community = str_to_title(community)
  )

#Note: Cleaned pollution sensor data and calculated average PM2.5 and NO2 values.
pollution_points <- pollution %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  group_by(sensor_name, latitude, longitude) %>%
  summarise(
    pm25 = mean(pm2_5ConcMass24HourMean.value, na.rm = TRUE),
    no2 = mean(no2Conc24HourMean.value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  #Note: Removed sensor points with missing PM2.5 or NO2 values.
  filter(!is.na(pm25), !is.na(no2)) %>%
  
  #Note: Created a standardized pollution burden score from PM2.5 and NO2.
  mutate(
    pollution_burden = as.numeric(scale(pm25)) + as.numeric(scale(no2))
  ) %>%
  
  #Note: Converted pollution sensor locations into spatial points.
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

#Note: Matched pollution sensor points to Chicago community areas.
pollution_area <- st_join(pollution_points, chicago_areas, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(area_numbe, community) %>%
  summarise(
    pm25 = mean(pm25, na.rm = TRUE),
    no2 = mean(no2, na.rm = TRUE),
    pollution_burden = mean(pollution_burden, na.rm = TRUE),
    .groups = "drop"
  )

#Note: Cleaned health data and selected asthma rate by community area.
health_clean <- health %>%
  mutate(
    community = str_to_title(Name),
    asthma_rate = `HCSATH_2023-2024`
  ) %>%
  select(community, asthma_rate)

#Note: Combined Chicago boundaries, pollution burden, and asthma data.
leaflet_data <- chicago_areas %>%
  left_join(pollution_area, by = c("area_numbe", "community")) %>%
  left_join(health_clean, by = "community") %>%
  
  #Note: Created a combined pollution and asthma burden score.
  mutate(
    combined_score = as.numeric(scale(pollution_burden)) + as.numeric(scale(asthma_rate)),
    
    #Note: Created popup text for interactive map labels.
    popup_text = paste0(
      "<b>", community, "</b>",
      "<br><b>PM2.5:</b> ", round(pm25, 2), " µg/m³",
      "<br><b>NO2:</b> ", round(no2, 2), " ppb",
      "<br><b>Pollution Burden:</b> ", round(pollution_burden, 2),
      "<br><b>Asthma Rate:</b> ", round(asthma_rate, 2),
      "<br><b>Pollution + Asthma Score:</b> ", round(combined_score, 2)
    )
  )

#Note: Created pink-white-green color palette for combined burden score.
pal <- colorNumeric(
  palette = c("#f8c8dc", "#ffffff", "#006d2c"),
  domain = leaflet_data$combined_score,
  na.color = "#bdbdbd"
)

#Note: Created interactive leaflet map with Chicago community polygons.
leaflet(leaflet_data) %>%
  
  #Note: Added light basemap underneath the community polygons.
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  #Note: Added community area polygons colored by pollution and asthma burden.
  addPolygons(
    fillColor = ~pal(combined_score),
    fillOpacity = 0.75,
    color = "#555555",
    weight = 0.7,
    popup = ~popup_text,
    label = ~paste0(community),
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#000000",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  
  #Note: Added legend for pollution and asthma burden.
  addLegend(
    pal = pal,
    values = ~combined_score,
    title = "Pollution + Asthma Burden",
    position = "bottomright"
  )
```
## Row

### Conclusion (asthma)

**Overall, I wanted to show the relationship between pollution and asthma levels. I aimed to explore weather asthma cases were higher based on traffic pollution and how these rates affect environmental burden. My first graph is a leaflet map that showcases asthma prevalence in the Chicagoland area. This figure showed some higher concentrations of asthma in the northern region, however these differences were not significant. I made this map to establish baseline levels of asthma prevalence throughout the city. The second figure was a mismatch graph made to see weather certain neighborhoods experience higher or lower expected levels of asthama based on traffic pollution. The findings revealed the same neighborhoods with higher levels of asthma on figure 1 were the same neighborhoods with more vulerability than expected: Lakeview, Austin, Albany Park. There was a slight positive correlation between asthma levels and pollution with some variability. Figure 3 is a scatter plot with dashed lines that indicate city-averages between traffic-related pollution and asthma prevalence. This figure shows weather asthma tends to increase as pollution increases to establish a baseline relationship between the two variables. Findings reveal the same outliers as previous figures have higher asthma levels than pollution levels. This suggests pollution does not fully explain asthma outcomes. Figure (4) attempts to uncover weather environmental justice predicts weather a neighborhood will have higher or lower asthma levels than expected. The findings reveal there is no strong pattern between the two variables as points are widely scattered with a few outliers. For example the neighborhood of Lakeview has low environmental justice burden but a high positive mismatch, the neighborhood of Austin has a high environmental burden, and high positive mismatch. The scatterplot (Figure 5) establishes the baseline relationship between pollution burden and asthma prevalence. It shows a slight positive trend, meaning asthma generally tends to increase with pollution, but the large spread and outliers demonstrate that pollution alone does not fully explain asthma outcomes across Chicago neighborhoods.The leaflet map (Figure 6) adds the spatial/environmental justice perspective. It reveals where combined pollution and asthma burdens cluster geographically. This helps identify neighborhoods experiencing disproportionately high burdens and supports the argument that health disparities are concentrated in certain areas.However, these random variations reveal no real pattern between the two. In conclusion, there is no real correlation between pollution and asthma levels. However, there are a few neighborhoods that are disproportionately affected for unknown reasons which could be potentially attributed to traffic, pm2.5, and NO2 pollution. The results suggest further analyses needs to be conducted in order to reveal asthma causation.** 

Poverty and Health
=======================================

## Row

```{r load-library, echo=FALSE, warning=FALSE, message=FALSE}
#Note: Data manipulation
library(tidyverse)
library(lubridate)

#Note: Interactive plots
library(plotly)
library(leaflet)

#Note: Mapping
library(sf)
library(tmap)

#Note: Statistical tools
library(corrplot)
library(broom)

#Note: Sunset color palette
sunset_colors <- c(
  "#C21807",  #Note: cherry red
  "#FF5E78",  #Note: pink
  "#FFA600",  #Note: orange
  "#FFD166"   #Note: yellow
)
```


```{r load-data, echo=FALSE, warning=FALSE, message=FALSE}
#Note: Environmental data\
env_url <- "https://raw.githubusercontent.com/fitzley/miniproj2/main/Open_Air_Chicago_Day_Aggregations_20260212(1)%20(3).csv"
env_data <- read.csv(env_url)

#Note: Biological health data
bio_data <- read.csv(url4)
```


```{r inspect-data, echo=FALSE, warning=FALSE, message=FALSE}
#Note: View structure of data sets
str(env_data)
str(bio_data)

#Note: View column names
colnames(env_data)
colnames(bio_data)
```


```{r clean-data-environmental, echo=FALSE, warning=FALSE, message=FALSE}

#Note: The environmental dataset contains repeated daily observations collected by air quality sensors across Chicago. Before analysis, the data must be simplified and standardized.

#Note:Goals of cleaning:

#Note:Keep only variables relevant to this project
#Note:Remove incomplete observations
#Note:Create neighborhood-level pollution summaries
#Note:Compare multiple pollution indicators together
#Note: Selected variables needed for analysis (pm25_24HourMean.value and bcFossilFuel24HourMean.value)


env_clean <- env_data %>%
  select(
    sensor_name,
    latitude,
    longitude,
    startofperiod,
    pm2_5ConcMass24HourMean.value,
    bcFossilFuel24HourMean.value
  ) %>%
  
#Note: Renamed variables for easier coding and readability
#Note: Shorter names make plotting and analysis cleaner
  
  rename(
    pm25 = pm2_5ConcMass24HourMean.value,
    fossil_fuel_bc = bcFossilFuel24HourMean.value
  )

#Note: Converted date column into proper date-time format
#Note: This prevents formatting problems later in the analysis

env_clean$startofperiod <- mdy_hms(env_clean$startofperiod)

#Note: Removed rows with missing pollution measurements
#Note: Missing values can distort averages and weaken relationships

env_clean <- env_clean %>%
  drop_na(pm25, fossil_fuel_bc)

#Note: Created long-term pollution averages by sensor location

sensor_avg <- env_clean %>%
  group_by(sensor_name, latitude, longitude) %>%
  summarise(
    avg_pm25 = mean(pm25, na.rm = TRUE),
    avg_fossil_bc = mean(fossil_fuel_bc, na.rm = TRUE)
  )

#Note: Preview cleaned environmental dataset

head(sensor_avg)
```

## Row {data-height=300}

### Why This Step Matters

PM2.5 and fossil fuel black carbon fluctuate daily due to weather conditions, traffic intensity, and seasonal changes. By averaging measurements across time, the analysis focuses on persistent exposure patterns instead of temporary fluctuations. Using two environmental indicators also strengthens the project because it helps determine whether multiple pollution variables support similar ecological conclusions.

Air pollution fluctuates daily due to weather, traffic, and seasonal conditions. By averaging pollution levels across time, the analysis focuses on persistent neighborhood exposure patterns instead of short-term variability. This creates a more biologically meaningful comparison with long-term social indicators like poverty.


```{r clean-data-biological }

#Note:The Chicago Health Atlas data set contains tract-level demographic and health information. 
#Note:This project focuses specifically on poverty because socioeconomic vulnerability often influences environmental exposure patterns.

#Note: Removed the first row if it contains metadata instead of observations
bio_clean <- bio_data[-1, ]

#Note: Selected only variables needed for this analysis
#Note: This keeps the workflow organized and easier to interpret

bio_clean <- bio_clean %>%
  select(
    Name,
    GEOID,
    Latitude,
    Longitude,
    `POV_2020.2024`
  )

#Note: Renamed columns for easier coding and readability
#Note: Simpler variable names make plotting and analysis cleaner

bio_clean <- bio_clean %>%
  rename(
    poverty_rate = `POV_2020.2024`,
    latitude = Latitude,
    longitude = Longitude
  )

#Note: Converted poverty values into numeric format
#Note: Public datasets often import numeric values as characters

bio_clean$poverty_rate <- as.numeric(bio_clean$poverty_rate)

#Note: Remove rows with missing poverty values
#Note: Missing values could distort statistical relationships

bio_clean <- bio_clean %>%
  drop_na(poverty_rate)

#Note: Preview cleaned biological dataset

head(bio_clean)

```

### Poverty as an Indicator

**Poverty was selected because it reflects more than income alone. In urban ecological studies, poverty is often connected to:**

reduced access to healthcare,
increased environmental exposure,
proximity to major roadways or industrial zones,
and higher long-term biological vulnerability.

This makes poverty a strong candidate for investigating whether environmental burdens are distributed unevenly across the city of Chicago.


```{r poverty_rate_map}

#Note: Create an interactive Chicago poverty vulnerability map.
#Note: This map establishes the spatial distribution of socioeconomic vulnerability
#Note: across Chicago census tracts.

#Note: Poverty was selected because it is frequently associated with:
#Note: environmental exposure, reduced healthcare access,
#Note: and increased long-term biological vulnerability.

#Note: Create custom sunset color variables
#Note: to maintain consistent project aesthetics.

sun_yellow <- "#FFD166"
sun_orange <- "#FFA600"
sun_pink <- "#FF5E78"
sun_red <- "#C21807"
dark_gray <- "#3A3A3A"

#Note: Convert coordinates into numeric values.

bio_clean$latitude <- as.numeric(bio_clean$latitude)
bio_clean$longitude <- as.numeric(bio_clean$longitude)

#Note: Remove rows with missing coordinates or poverty values.

poverty_map_df <- bio_clean %>%
  filter(
    !is.na(latitude),
    !is.na(longitude),
    !is.na(poverty_rate)
  )

#Note: Create interactive poverty map using Plotly Mapbox.
#Note: Larger and darker markers indicate higher poverty vulnerability.

plot_ly(
  poverty_map_df,
  
  type = "scattermapbox",
  
  #Note: Geographic coordinates
  lat = ~latitude,
  lon = ~longitude,
  
  mode = "markers",
  
  #Note: Marker styling
  marker = list(
    
    #Note: Marker size based on poverty rate
    size = poverty_map_df$poverty_rate * 0.7,
    
    #Note: Marker color based on poverty rate
    color = poverty_map_df$poverty_rate,
    
    #Note: Sunset themed color gradient
    colorscale = list(
      c(0, sun_yellow),
      c(0.35, sun_orange),
      c(0.7, sun_pink),
      c(1, sun_red)
    ),
    
    showscale = TRUE,
    
    #Note: Marker border styling
    line = list(
      color = dark_gray,
      width = 1
    )
  ),
  
  #Note: Hover information
  text = ~paste(
    "<b>Census Tract:</b>", Name,
    "<br><b>Poverty Rate:</b>", round(poverty_rate, 2)
  ),
  
  hoverinfo = "text"
  
) %>%
  
  layout(
    
    title = "Interactive Chicago Poverty Vulnerability Map",
    
    #Note: Chicago-centered map
    mapbox = list(
      style = "carto-positron",
      zoom = 9,
      center = list(
        lat = 41.85,
        lon = -87.68
      )
    ),
    
    margin = list(l = 0, r = 0, b = 0, t = 50)
  )

```

## Row

### Analysis of Interactive Poverty Vulnerability Map

The poverty vulnerability map demonstrates that socioeconomic stress is not evenly distributed across Chicago neighborhoods. Several census tracts display larger and darker sunset-colored markers, indicating areas experiencing elevated poverty rates compared to surrounding communities. These spatial clustering patterns suggest that social vulnerability may concentrate within particular regions of the city rather than occurring randomly across census tracts. From an ecological perspective, this is important because poverty is often associated with reduced healthcare access, environmental exposure, and increased long-term biological stress. Overall, the visualization supports the project’s hypothesis that vulnerable populations may experience overlapping environmental and socioeconomic burdens throughout Chicago.


```{r poverty_distribution_map}

#Note: Create an interactive poverty distribution intensity map.
#Note: This visualization focuses on identifying clusters
#Note: of elevated poverty vulnerability throughout Chicago.

#Note: The purpose of this plot is to determine whether
#Note: socioeconomic burden appears concentrated geographically.

#Note: Create grouped poverty categories.
#Note: Grouping improves interpretation of neighborhood-level vulnerability.

poverty_distribution_df <- poverty_map_df %>%
  
  mutate(
    poverty_group = case_when(
      poverty_rate < 10 ~ "Lower Vulnerability",
      poverty_rate >= 10 & poverty_rate < 20 ~ "Moderate Vulnerability",
      poverty_rate >= 20 ~ "Higher Vulnerability"
    )
  )

#Note: Create interactive distribution map.

plot_ly(
  poverty_distribution_df,
  
  type = "scattermapbox",
  
  lat = ~latitude,
  lon = ~longitude,
  
  mode = "markers",
  
  #Note: Marker styling by grouped poverty vulnerability
  marker = list(
    
    size = 12,
    
    color = ~poverty_rate,
    
    colorscale = list(
      c(0, sun_yellow),
      c(0.35, sun_orange),
      c(0.7, sun_pink),
      c(1, sun_red)
    ),
    
    opacity = 0.8,
    
    line = list(
      color = dark_gray,
      width = 1
    ),
    
    showscale = TRUE
  ),
  
  #Note: Hover information
  text = ~paste(
    "<b>Census Tract:</b>", Name,
    "<br><b>Poverty Rate:</b>", round(poverty_rate, 2),
    "<br><b>Vulnerability Group:</b>", poverty_group
  ),
  
  hoverinfo = "text"
  
) %>%
  
  layout(
    
    title = "Chicago Poverty Vulnerability Distribution",
    
    mapbox = list(
      style = "carto-positron",
      zoom = 9,
      center = list(
        lat = 41.85,
        lon = -87.68
      )
    ),
    
    margin = list(l = 0, r = 0, b = 0, t = 50)
  )

```


### Analysis of Interactive Poverty Distribution Map

The poverty distribution map highlights strong geographic clustering patterns across Chicago census tracts. Higher vulnerability neighborhoods appear concentrated within specific areas, while lower poverty regions remain more spatially separated throughout the city. The darker sunset-colored markers emphasize how socioeconomic burden may accumulate within particular communities instead of being evenly dispersed across urban space. These patterns are ecologically significant because concentrated poverty may increase vulnerability to environmental exposure, limited healthcare access, and long-term biological stressors. Overall, the visualization supports the project’s broader conclusion that social vulnerability patterns appear spatially structured and may overlap with environmental inequality throughout Chicago.


Diabetes
=====================================

## Row {data-height="450"} 

### Introduction

Diabetes prevalence varies across Chicago community areas, suggesting that biological health outcomes are not evenly distributed across space. In Mini Project 2, I examined diabetes as one of the major neighborhood-level health outcomes. For Mini Project 3, I will examine whether traffic related environmental exposure risk aligns with diabetes prevalence across Chicago communities. 

The goal of this analysis is not to argue that traffic environmental exposure directly causes diabetes. Instead, traffic risk is treated as environmental exposure indicator that may reflect broader patterns of neighborhood vulnerability, environmental burden, and health inequity. 

**Research Question:**

How well does traffic-related environmental risk align with diabetes prevalence across Chicago community areas?

**Hypothesis:**

Community areas with higher environmental exposure may show higher diabetes prevalence, but the relationship is expected to be imperfect because diabetes is influenced by many social, economic, political, and environmental factors beyond traffic exposure alone. 

### Methods

The analysis uses datasets from the group GitHub repository. The Open Air Chicago dataset from Mini Project 1 provides environmental context through PM2.5 sensor observations, while the Chicago Health Atlas dataset from Mini Project 2 provides community-area variables, including traffic risk and diabetes prevalence.

Because the Open Air Chicago data are measured at the sensor/day level and diabetes prevalence is reported at the community-area level, the datasets do not align perfectly to scale. To make the comparison meaningful at the neighborhood level, this analysis uses traffic-related environmental risk ('TRF_2020') from the Chicago Health Atlas dataset as the main community-level environmental predictor.

Diabetes burden is represented by ('HCSDIA_2023.2024'), which was treated as the estimated number of adults with diabetes. To make this comparable across community areas, diabetes prevalence was calculated by dividing this value by the total population and multiplying by 100. Traffic risk and diabetes prevalence are compared using a scatterplot, correlation test, and linear regression model. A diabetes mismatch index is then calculated by subtracting predicted diabetes prevalence from observed diabetes prevalence. Positive mismatch values indicate communities with higher diabetes prevalence than expected based on traffic risk, while negative values indicate communities with lower diabetes prevalence than expected. 

## Row {data-height="450"}

### Result 1: Prepare Diabetes and Traffic Data
```{r diabetes-traffic-setup, include=FALSE}
diabetes_traffic <- chicago_map %>%
  st_drop_geometry() %>%
  filter(
    !is.na(TRF_2020),
    !is.na(diabetes_rate)
  ) %>%
  mutate(
    traffic_risk = TRF_2020,
    diabetes_prevalence = diabetes_rate
  )

traffic_model <- lm(
  diabetes_prevalence ~ traffic_risk,
  data = diabetes_traffic
)

traffic_cor <- cor.test(
  diabetes_traffic$traffic_risk,
  diabetes_traffic$diabetes_prevalence
)

traffic_model_results <- broom::tidy(
  traffic_model,
  conf.int = TRUE
)
```

```{r diabetes-preview-table, echo=FALSE, warning=FALSE, message=FALSE}
diabetes_preview <- diabetes_traffic %>%
  select(
    Community = community,
    `Traffic Risk` = traffic_risk,
    `Diabetes Prevalence (%)` = diabetes_prevalence
  ) %>%
  arrange(desc(`Diabetes Prevalence (%)`)) %>%
  mutate(
    `Traffic Risk` = round(`Traffic Risk`, 2),
    `Diabetes Prevalence (%)` = round(`Diabetes Prevalence (%)`, 2)
  )

DT::datatable(
  diabetes_preview,
  caption = "Prepared Traffic Risk and Diabetes Prevalence Data",
  options = list(
    pageLength = 10,
    scrollX = TRUE
  )
)
```

## Row {data-height="500"}

### Result 2: Traffic Risk vs Diabetes Prevalence
```{r traffic-diabetes-plot, echo=FALSE, warning=FALSE, message=FALSE}
p1 <- ggplot(
  diabetes_traffic,
  aes(
    x = traffic_risk,
    y = diabetes_prevalence,
    text = paste0(
      "<b>", community, "</b><br>",
      "Traffic Risk: ", round(traffic_risk, 2), "<br>",
      "Diabetes Prevalence: ", round(diabetes_prevalence, 2), "%"
    )
  )
) +
  geom_point(color = "#C77DFF", size = 3, alpha = 0.75) + 
  geom_smooth(
    method = "lm",
    color = "#7B2CBF",
    se = TRUE
  ) +
  scale_x_log10() +
  scale_y_continuous(labels = function(x) paste0(round(x, 1), "%")) +
  labs(
    title = "Traffic Risk and Diabetes Prevalence",
    subtitle = "Testing whether traffic-related environmental burden aligns with diabetes patterns",
    x = "Traffic Risk (log scale)",
    y = "Diabetes Prevalence (% Adults)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(color = "black", face = "bold"),
    plot.subtitle = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.text = element_text(color = "black")
  )

ggplotly(p1, tooltip = "text") %>%
  layout(
    autosize = TRUE,
    margin = list(l = 70, r = 40, b = 70, t = 90)
  ) %>%
  config(displayModeBar = FALSE, responsive = TRUE)
```

## Row {data-height="300"}

### Result 3: Correlation Test
```{r correlation-test, echo=FALSE, warning=FALSE, message=FALSE}
cor_table <- tibble(
  relationship = "Traffic Risk vs Diabetes Prevalence",
  correlation = round(as.numeric(traffic_cor$estimate), 3),
  p_value = round(traffic_cor$p.value, 4)
)

DT::datatable(
  cor_table,
  caption = "Correlation Between Traffic Risk and Diabetes Prevalence",
  options = list(pageLength = 5)
)
```

### Result 4: Regression Model
```{r regression-model, echo=FALSE, warning=FALSE, message=FALSE}
DT::datatable(
  traffic_model_results,
  caption = "Regression Model Predicting Diabetes Prevalence from Traffic Risk",
  options = list(pageLength = 5)
)
```

## Row {data-height="650"}

### Result 5: Diabetes Mismatch Index
```{r mismatch-index, include=FALSE}
diabetes_traffic <- diabetes_traffic %>%
  mutate(
    expected_diabetes = predict(traffic_model, newdata = diabetes_traffic),
    diabetes_mismatch = diabetes_prevalence - expected_diabetes,
    mismatch_category = case_when(
      diabetes_mismatch > 0 ~ "Higher diabetes than expected",
      diabetes_mismatch < 0 ~ "Lower diabetes than expected",
      TRUE ~ "About as expected"
    ),
    hover_text = paste0(
      "<b>", community, "</b><br>",
      "Traffic Risk: ", round(traffic_risk, 2), "<br>",
      "Observed Diabetes: ", round(diabetes_prevalence, 2), "%<br>",
      "Expected Diabetes: ", round(expected_diabetes, 2), "%<br>",
      "Mismatch: ", round(diabetes_mismatch, 2), "<br>",
      "Category: ", mismatch_category
    )
  )
```

```{r mismatch-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.width=11, fig.height=6, out.width="100%"}
p2 <- ggplot(
  diabetes_traffic,
  aes(
    x = expected_diabetes,
    y = diabetes_prevalence,
    color = diabetes_mismatch,
    text = hover_text
  )
) +
  geom_point(size = 4, alpha = 0.85) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "gray40"
  ) +
  scale_color_gradient2(
    low = "#FF4FA3",
    mid = "#C77DFF",
    high = "#3A0CA3",
    midpoint = 0,
    name = "Mismatch"
  ) +
  labs(
    title = "Diabetes Mismatch Index",
    subtitle = "Observed diabetes compared with diabetes predicted from traffic risk",
    x = "Expected Diabetes Prevalence (%)",
    y = "Observed Diabetes Prevalence (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(color = "black", face = "bold"),
    plot.subtitle = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(color = "black"),
    legend.text = element_text(color = "black")
  )

ggplotly(p2, tooltip = "text") %>%
  layout(
    autosize = TRUE,
    margin = list(l = 80, r = 40, b = 80, t = 90)
  ) %>%
  config(displayModeBar = FALSE, responsive = TRUE)
```

## Row {data-height="850"}

### Result 6: Communities With the Strongest Diabetes Mismatch
```{r strongest-mismatch, echo=FALSE, warning=FALSE, message=FALSE, fig.width=11, fig.height=8, out.width="100%"}
mismatch_ranked <- diabetes_traffic %>%
  arrange(desc(abs(diabetes_mismatch))) %>%
  mutate(community = factor(community, levels = rev(unique(community))))

p3 <- ggplot(
  mismatch_ranked,
  aes(
    x = diabetes_mismatch,
    y = community,
    fill = mismatch_category,
    text = hover_text
  )
) +
  geom_col(width = 0.55) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "Higher diabetes than expected" = "#3A0CA3",
      "Lower diabetes than expected" = "#FF4FA3",
      "About as expected" = "gray70"
    )
  ) +
  labs(
    title = "Communities With the Strongest Diabetes Mismatch",
    subtitle = "Positive values indicate higher diabetes than predicted by traffic risk",
    x = "Diabetes Mismatch",
    y = NULL,
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(color = "black", face = "bold"),
    plot.subtitle = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    axis.text.y = element_text(size = 8),
    legend.title = element_text(color = "black"),
    legend.text = element_text(color = "black"),
    legend.position = "bottom"
  )

ggplotly(p3, tooltip = "text") %>%
  layout(
    autosize = TRUE,
    margin = list(l = 180, r = 40, b = 80, t = 90)
  ) %>%
  config(displayModeBar = FALSE, responsive = TRUE)
```

## Row {data-height="420"}

### Analysis/Discussion

The comparison between traffic-related risk and diabetes prevalence shows a weak and inconsistent relationship across Chicago community areas. Although the original expectation was that higher traffic risk would align with higher diabetes prevalence, the scatterplot does not show a clear upward trend. Instead, communities are widely dispersed, meaning that similar levels of traffic risk can be associated with very different diabetes prevalence values. This indicates that traffic risk alone does not fully explain diabetes patterns at the community-area scale.

The linear regression model supported the same interpretation. Traffic risk alone does not meaningfully explain the observed variation in diabetes prevalence. This is useful because it helps avoid overstating the relationship or treating the association as causal when the statistical evidence is weak.

The diabetes mismatch index adds more depth to the analysis by showing where observed diabetes prevalence differs from what would be expected based on traffic risk. Communities with positive mismatch values had higher diabetes prevalence than predicted, while communities with negative mismatch values had lower diabetes prevalence than predicted. This makes the analysis stronger because it identifies where the expected relationship breaks down, rather than only asking whether traffic risk predicts diabetes overall.

These deviations suggest that diabetes prevalence is shaped by more than one environmental factor. While traffic risk still provides useful environmental context, it does not capture the full range of conditions influencing diabetes patterns. Factors such as accessibility to food and healthcare, income, age structure, neighborhood resources, and broader inequalities may help explain why some communities deviate strongly from the prediction.

## Row {data-height="320"}

### Limitations

One limitation of this analysis is the scale mismatch. The Open Air Chicago dataset is measured at the sensor/day level, while the Chicago Health Atlas dataset is reported at the community-area level. Because of this, the Open Air Chicago dataset is used as a means to establish environmental context, while the 'TRF_2020' variable from the Chicago Health Atlas dataset is used as a community-level environmental predictor for the diabetes analysis.  

A second limitation is the time mismatch. Traffic risk is from 2020, while diabetes prevalence is from 2023-2024. Because these variables were not measured during the exact same period, the results should be interpreted as explanatory spatial associations.

One final limitation, perhaps the most important, is that traffic risk only represents one environmental variable. Diabetes prevalence is largely influenced by many socioeconomic, political, and structural factors that cannot be captured by traffic risk alone.  

## Row {data-height="260"}

### Conclusions
 
Overall, this analysis shows that traffic-related environmental risk alone does not strongly explain diabetes prevalence across Chicago communities. The scatterplot, correlation test, and regression model suggest that the relationship is weak, meaning that higher traffic risk does not consistently predict diabetes prevalence.

However, the mismatch index strengthens the analysis by showing where the expected relationship breaks down. Some communities have higher prevalence based on risk, while others experience the opposite. This supports the idea that environmental and biological datasets can be meaningfully compared, but their relationships cannot be explained by one variable alone.

## Row {data-height="300"}

### References

McAlexander, T. P., Ryan, V., Uddin, J., Kanchi, R., Thorpe, L., Schwartz, B. S., Carson, A., Rolka, D. B., Adhikari, S., Pollak, J., Lopez, P., Smith, M., Meeker, M., & McClure, L. A. (2023). Associations between PM2.5 and O3 exposures and new onset type 2 diabetes in regional and national samples in the United States. *Environmental Research, 239*, 1-9. https://doi.org/10.1016/j.envres.2023.117248

Montressor-López, J. A., Reading, S. R., Yanosky, J. D., Mittleman, M. A., Bell, R. A., Crume, T. L., Dabelea, D., Dolan, L., D'Agostino Jr., R. B., Marcovina, S. M., Pihoker, C., Reynolds, K., Urbina, E., Liese, A. D., Quirós-Alcalá, L., Smith, J. C., de Mesquita, P. J. B., & Puett, R.C. (2021). The relationship between traffic-related air pollution exposures and allostatic load among young with type 1 diabetes in the SEARCH cohort. *Environmental Research, 197*, 1-7. https://doi.org/10.1016/j.envres.2021.111075


Conclusions
=====================================

## Row 

This project investigated whether neighborhoods in Chicago with higher socioeconomic vulnerability also experience greater exposure to air pollution, specifically black carbon concentrations. By combining environmental sensor observations with Chicago Health Atlas indicators,the analysis explored whether environmental burden and poverty overlap in a way that reflects realistic urban ecological conditions rather than random statistical coincidence.  
Overall, the findings suggest that there is a meaningful relationship between poverty and air pollution exposure across Chicago census tracts. Neighborhoods with higher poverty rates frequently aligned with higher average black carbon concentrations, indicating that environmental stressors may disproportionately affect socially vulnerable communities.  
These findings support broader environmental justice research showing that low-income populations are often located closer to major traffic corridors, industrial infrastructure, and other combustion-related pollution sources.  
The relationship observed throughout the analysis also demonstrated ecological credibility. Black carbon pollution is strongly associated with traffic emissions and combustion activity, making it biologically and environmentally reasonable that communities with fewer economic resources may experience greater exposure. The consistency of these patterns across visualizations and spatial comparisons suggests that the relationship reflects broader structural urban inequalities rather than isolated statistical patterns.  
However, this analysis also has important limitations. One major limitation was the mismatch between the two datasets. The environmental sensor data represented long-term continuous observations collected over time, while the Chicago Health Atlas indicators were census tract snapshots collected during a different period. Because the datasets were not perfectly synchronized temporally, the analysis may not fully capture changes in neighborhood conditions over time.  
Another limitation involved differences in spatial scale and data coverage. Environmental sensors were not evenly distributed across all Chicago neighborhoods, meaning some census tracts may have been represented more accurately than others. In addition, averaging
environmental measurements across census tracts may have reduced local variability and hidden smaller pollution hotspots within neighborhoods.  
As a group, we also encountered challenges involving data cleaning, standardization, and dataset integration. Combining two independent datasets required extensive preprocessing to ensure that variables could be compared responsibly. Differences in formatting, geographic boundaries, missing values, and measurement units introduced additional uncertainty into the analysis process. Because this project relied primarily on observational data, there were also limits to how strongly we could interpret causal relationships between poverty and pollution.  
Time constraints and limited data availability also affected the scope of the project. While this analysis focused specifically on black carbon and poverty rates, incorporating additional variables such as particulate matter, asthma prevalence, traffic density, heat exposure, or healthcare access could provide a more comprehensive understanding of environmental vulnerability in Chicago. Future research could also benefit from longitudinal analysis and stronger spatial modeling techniques.  
Despite these limitations, the project successfully demonstrated how environmental and biological datasets can be combined to investigate urban environmental inequality in a scientifically responsible way. The results support the conclusion that environmental burdens are not distributed equally across Chicago neighborhoods and that socioeconomic vulnerability may be associated with higher exposure to pollution. Most importantly, this project highlights the importance of combining environmental science, biology, and public health perspectives to better understand how ecological conditions and social inequality intersect in urban communities.


Acknowledgements
=====================================

<div style="background-color:#6A0DAD; padding:20px; border-radius:20px; margin-bottom:20px;">

## 
Maddie Cassagrande

- Conducted in-depth analysis of diabetes burden in relation to environmental exposure, creating the diabetes section.
- Helped generate major project ideas  
- Assisted with data filtering and preparation
- Colors: Purple, Bright pink

</div>

<div style="background-color:#fd9040; padding:20px; border-radius:12px; margin-bottom:20px;">

## 
Alehna Crowe

- Conducted in-depth analysis of overall disease trends and obesity prevalence in vulnerable populations  
- Developed the Disease Overview and Obesity sections  
- Combined and organized the majority of the project datasets  
- Helped create the flexdashboard template  
- Assisted with brainstorming and development of the final project structure  
- Created the shared GitHub workspace  
- Assisted with data filtering and preparation 
- Colors:<div style="display:flex; gap:8px; align-items:center; margin:6px 0;">
  <div style="background:#48bea1; width:40px; height:2.4em; border-radius:3px;"></div>
  <div style="background:#f4bb8f; width:40px; height:2.4em; border-radius:3px;"></div>
  <div style="background:#7b4419; width:40px; height:2.4em; border-radius:3px;"></div>
  <div style="background:#3f6c11; width:40px; height:2.4em; border-radius:3px;"></div>
  <div style="background:#ca5603; width:40px; height:2.4em; border-radius:3px;"></div>
  <div style="background:#cccccc; width:40px; height:2.4em; border-radius:3px;"></div>
  <div style="background:#ef0300; width:40px; height:2.4em; border-radius:3px;"></div>
  <div style="background:#2e4f4f; width:40px; height:2.4em; border-radius:3px;"></div>
  
</div> 

<div style="background-color:#92d895; padding:20px; border-radius:12px; margin-bottom:20px;">

## 
Keyla Estrada

- Conducted in-depth analysis of pollution trends and asthma prevalence in relation to pollution  
- Developed the Pollution Overview and Asthma sections  
- Helped organize and integrate the flexdashboard structure  
- Created the acknowledgements section  
- Maintained and integrated class notes into the project workflow  
- Assisted with data integration and filtering
- Created navigation dashboard
- Colors: Green, Dark-light Pink
</div>

<div style="background-color:#ffc425; padding:20px; border-radius:12px; margin-bottom:20px;">

## 
Lizbeth Macias

- Conducted in-depth analysis of hypertension prevalence in relation to pollution  
- Developed the Hypertension section  
- Helped combine and organize datasets  
- Assisted in creating the flexdashboard template  
- Contributed to brainstorming and project planning  
- Created the shared Google document for collaborative planning  
- Developed the abstract/introduction and conclusion sections  
- Assisted with data filtering and preparation 
- Created designated branch on Github
- Colors: Orange, Yellow
</div>